Machine learning approaches identify immunologic signatures of total and intact HIV DNA during long-term antiretroviral therapy

Version 1

Abstract

Abstract

Antiretroviral therapy (ART) halts HIV replication; however, cellular / immue cell viral reservoirs persist despite ART. Understanding the interplay between the HIV reservoir, immune perturbations, and HIV-specific immune responses on ART may yield insights into HIV persistence. A cross-sectional study of peripheral blood samples from 115 people with HIV (PWH) on long-term ART was conducted. High-dimensional immunophenotyping, quantification of HIV-specific T cell responses, and the intact proviral DNA assay (IPDA) were performed. Total and intact HIV DNA was positively correlated with T cell activation and exhaustion. Years of ART and select bifunctional HIV-specific CD4 T cell responses were negatively correlated with the percentage of intact proviruses. A Leave-One-Covariate-Out (LOCO) inference approach identified specific HIV reservoir and clinical-demographic parameters that were particularly important in predicting select immunophenotypes. Dimension reduction revealed two main clusters of PWH with distinct reservoirs. Additionally, machine learning approaches identified specific combinations of immune and clinical-demographic parameters that predicted with approximately 70% accuracy whether a given participant had qualitatively high or low levels of total or intact HIV DNA. The techniques described here may be useful for assessing global patterns within the increasingly high-dimensional data used in HIV reservoir and other studies of complex biology.

Introduction

The development of ART has vastly improved morbidity and mortality associated with HIV infection. However, a long-lived proviral reservoir precludes a cure of infection (1–4). Furthermore, despite ART, PWH are still at increased risk of morbidity and mortality related to sequelae of persistent immune activation. This persistent immune activation is thought to be multifactorial: a product of residual immune pertubation from untreated HIV infection, persistent proviral expression during ART, and specific clinical risk factors for immune activation that are enriched amongst PWH. Understanding the interplay between the HIV reservoir and persistent immune activation during long-term ART is important to inform strategies for a cure.

While a number of cell types have been proposed as contributing to the HIV reservoir, the largest and most well-characterized reservoir is within CD4 T cells. Molecular analyses of residual HIV proviruses in CD4 T cells during ART have shown that the reservoir is comprised of a heterogeneous mixture of full-length replication-competent proviruses and a larger set of defective proviruses containing internal deletions or APOBEC-mediated hypermutation(s) (5–7). Infected cells can undergo clonal expansion and contraction over time, potentially in response to homeostatic cues or antigen-driven activation (8–10). Studies of SIV-infected non-human primates and HIV-infected humans show that a pool of latently-infected cells forms very early during infection (3,11). However, studies suggest the reservoir is enriched in viral variants that were circulating at the time of ART initiation, indicating that initiation of ART may promote the entry of recently infected cells into the reservoir (12, 13).

HIV infected cells are rare in peripheral blood, making measuring and characterizing the reservoir challenging. Additionally, the size of the reservoir is highly variable between individuals, ranging over multiple logs in scale. During therapy, a subset of the reservoir remains transcriptionally active, expressing viral RNA and, in some cases, viral protein, leading to persistent viremia or “blips” during therapy (14). The factors that regulate reservoir size and transcriptional activity are only beginning to become clear. Similarly, the impact of the latent reservoir on the immune system of PWH is not fully understood. Over time on therapy, intact proviruses become preferentially depleted from transcriptionally active genomic regions and enriched in heterochromatic regions, suggesting that ongoing viral expression promotes infected cell depletion, either through a cytopathic effect from viral proteins or from immune killing of infected cells with residual expression (15).

In addition to the presence of a latent HIV reservoir, PWH on therapy exhibit a persistent level of T cell immune activation and other altered phenotypes compared to uninfected individuals, involving cells of both the innate and adaptive immune systems. Specifically, HIV infection is associated with increased levels of inflammatory biomarkers and increased levels of activation for immune cells (16–19). PWH also exhibit a depleted naïve T cell compartment (20) and increased expression of activation/exhaustion and senescence markers such as PD-1 and KLRG1 (21–24). These alterations may contribute to increased levels of disease and death in PWH. The molecular mechanisms underlying persistent immune activation and dysfunction for PWH on ART are unclear, but it is possible that ongoing viral gene expression and reverse transcription generates ligands for pattern recognition receptors in the innate immune system (25). Viral protein expression during therapy likely maintains antigen-specific cells of the adaptive immune system. Given that viral protein expression may be relatively rare during therapy, another important potential cause of persistent immune activation during ART is persistent sequelae of the massive disruptions to T cell homeostasis and lymphoid structures that occur during untreated HIV infection.

It is possible that the HIV reservoir and the host immune system interact in multiple complex ways during therapy. The dynamics of how the HIV reservoir affects the immune system and vice versa are important to characterize and may lead to novel ideas to promote reservoir depletion or to inhibit immune dysfunction that may be related to ongoing HIV expression in persistently infected cells. A number of previous studies have examined the relationship between reservoir size and the immune system (26–45). Consistent major findings from these studies include enrichment of the HIV reservoir in CD4 T cells expressing activation and exhaustion markers, positive correlation of the reservoir with central memory T cells (Tcm) and transitional memory (Ttm) T cell subsets, and inverse correlation of the HIV reservoir with the CD4 nadir and the CD4/CD8 T cell ratio. One important limitation of most of these studies is that they were not able to distinguish intact proviruses from defective proviruses, making the relationship between the immune system and the intact reservoir unclear. Recent studies indicate enrichment of the intact HIV reservoir in cells expressing VLA-4, but no correlation with classic T cell activation markers (26, 37). Additional assessment of immune correlates of the intact HIV reservoir in additional cohorts is needed. Furthermore, most of these studies examined a limited number of surface markers and may have missed associations that would be revealed by a higher-resolution examination of immunophenotypes. Finally, there is a recent, growing appreciation for the profound influence of ART initiation timing and length of treatment on the size and composition of the HIV reservoir that must be accounted for in assessing persistent immune activation in PWH on therapy (46, 47).

In this study, we use a mixture of traditional statistical approaches and novel data science methods to assess the relationship between the immune system and the HIV reservoir across a cohort of 115 PWH. Collectively, this study highlights the utility of machine learning approaches to identify otherwise imperceptible global patterns in high-parameter studies of the HIV reservoir. Additionally, these approaches corroborate recent findings in the HIV persistence field regarding selective intact proviral decay and immune dynamics and highlight the complex immunologic signatures of HIV latency.

Results

Cohort description, reservoir quantification, and immunophenotyping

115 participants with HIV infection for at least one year and who had been receiving antiretroviral therapy (ART) for at least 0.9 years (median 9 years, IQR 5.2-16.6) were studied (Table 1). The study cohort was 77 percent male and median participant age was 45.

The Intact Proviral DNA assay (IPDA) was performed on isolated total CD4 T cells (48). The IPDA enables estimation of the frequency of total proviruses, intact proviruses, and the percent intact of total proviruses for each participant. Immunophenotyping was performed using 25-color spectral flow cytometry on peripheral blood mononuclear cells (PBMCs) from each study participant. A representative flow gating is shown in Figure S1. Briefly, live lymphocytes were first defined as T cells (CD3+/CD56-), NK cells (CD3-/CD56+), or NKT cells (CD3+/CD56+). T cells were then further subdivided as CD4+ or CD8+ T cells, and then as naïve T cells (Tn, CD45RA+/CCR7+), central memory T cells (Tcm, CD45RA-/CCR7+), effector memory cells (Tem, CD45RA-/CCR7-) and terminally differentiated effector cells (Teff, CD45RA+/CCR7-). Expression of various surface markers within each of these subsets was then examined (CD38, HLA-DR, PD-1, KLRG1, CD127, CD27, NKG2A). In addition to measuring the baseline abundance of surface proteins, we also stimulated PBMCs from each participant with a pool of HIV-derived peptides and measured intracellular cytokines TNFα, IL-2, and IFNγ as well as the surface expression of the degranulation marker CD107a following background subtraction of the signal from unstimulated cells. Frequencies of CD4 or CD8 T cells that were single- or double-positive for TNFα, IL-2, IFNγ, or CD107 were used for analysis.

These analyses resulted in the determination of a total of 144 parameters for the cohort of n = 115 PWH. These 144 parameters include 133 immunophenotypic cell frequencies, 3 HIV reservoir parameters (frequency of intact HIV DNA, frequency of total HIV DNA, and percentage intact of total HIV DNA), and 8 clinical-demographic parameters. Clinical-demographic variables (Table 1) included age, biological sex, years of ART, estimated years of HIV infection prior to ART (here, we use categorical variables: years before ART=NA, years before ART< 1, years before ART≥ 1), CD4 nadir, most recent CD4 T cell count, and race (Caucasian, African-American, other). The median and interquartile range (IQR) for all immune cells population variables are presented in Table S2. No significant differences in the reservoir based on biological sex or race were observed (Table S1).

Correlation of total, intact, and percent intact HIV DNA with immunophenotypes

For the initial analysis, we performed Spearman correlations of the clinical-demographic and immunophenotypic data with HIV reservoir metrics. 69 host variables correlated (unadjusted p < 0.05) with one or more characteristics of the HIV reservoir (Table S3). Following Benjamini-Hochberg p-value adjustment for multiple comparisons using a false discovery rate correction at 5%, 31 variables were found to be correlated with total HIV DNA, 3 with intact HIV DNA, and 4 with the percentage of intact HIV DNA (Table 2). Overall, this analysis indicated that the HIV reservoir size is significantly associated with the abundance of several immunophenotype frequencies and clinical-demographic variables. Of note, the correlations observed were generally weak to moderate in magnitude (the mean of the absolute values of Spearman correlation coefficients was 0.31 for total HIV DNA, 0.33 for intact HIV DNA, and 0.35 for percent intact HIV DNA).

With regard to clinical-demographic variables, CD4 nadir (Spearman r = −0.32) was negatively correlated with total HIV DNA, whereas age (r = 0.31) and years of ART (r = 0.31) were positively correlated with total reservoir frequencies (Table 2). In contrast, years of ART (r = −0.45) was negatively correlated with the percentage of intact proviruses (Table 2). Given the growing appreciation in the field for the role of ART initiation and timing on the HIV reservoir and immune recovery, we examined this association further (Figure 1). Correlation plots of total, intact, and percentage intact HIV DNA versus years of ART were visualized (Figure 1A-C). As expected, the plots showed no significant correlation for intact HIV DNA versus years of ART (Figure 1B), while total reservoir size was positively correlated with the time of ART (Figure 1A, Spearman r = 0.31). Interestingly, when we examined a plot of percent intact proviruses versus time on therapy (Figure 1C), we observed a biphasic decay pattern. Indeed, when we fitted piece-wise linear functions with allowances for up to 2 breaks, the best-fit model had decay slope changes at 1.6 and 9 years of ART (R2 score of piece-wise linear model with two breaks was 44.18%, while linear model without any breaks – 16.3%). In PWH within approximately 0-6 years of ART initiation, a significant proportion of the participants exhibited high fractions of intact proviruses (50-90%). After approximately 6 years of ART, however, the majority of participants had reservoirs for which intact proviruses represented a minor fraction of the overall reservoir. Of note, several immune features also demonstrated statistically significant correlations with years of ART, including CD8 T cell expression of CD38, CD8 Tcm frequencies, and CD107a+IL-2+IFNγ-TNFα-CD8 T cells (Figures S2, S3, S4, Table S4).

Duration of treatment and the HIV reservoir.

Scatterplots for years of ART vs HIV reservoir size total reservoir size (A), intact reservoir size (B) and percent intact (C) are shown. Each dot represents an individual study participant. Correlation coefficients and corresponding p-values are shown for each plot. Patients that have missing values of years of ART were not included in the plot. For percent intact, piece-wise linear function with two breaks is fitted. For total reservoir size the linear function is fitted.

Examination of immunophenotypic parameter correlations with HIV DNA metrics demonstrated several notable correlations (Table 2, Figures S2, S3, S4). As previously reported, CD4 T cell frequencies were negatively correlated with total HIV DNA frequencies (r = −0.36), whereas CD8 T cells were positively correlated (r = 0.41). This relationship held true for intact HIV DNA frequencies but not for the percentage of intact HIV DNA (Table 2). Frequencies of CD4 and CD8 T naïve cells, as well as their expression of CD38, were significantly negatively correlated with total HIV DNA frequency. By contrast, CD4 and CD8 Tcm and Tem cells were weakly positively correlated with total HIV DNA. The direction of these correlations was concordant for intact HIV DNA frequencies, albeit without statistical significance. Markers of T cell activation and exhaustion were weakly but significantly positively correlated with total HIV DNA, including NKG2A+ CD4 T, PD-1+ CD4 T, HLA-DR+ CD4 T, PD-1+ CD8 T, and PD-1+/CCR7+CD8 T frequencies, among others (Table 2, Figures S2, S3, S4). Again, the direction of these correlations were conserved for intact HIV DNA frequencies, albeit without statistical significance. For both intact and total HIV DNA, an inverse correlation with the CD4/CD8 T cell ratio was observed (Figure S5).

In general, the magnitude of correlation was stronger for total HIV DNA compared to intact HIV DNA (Table 2). This pattern may reflect a larger influence of total HIV DNA on immunophenotype of the host. A notable exception to this pattern was CD127+ CD4 T cell frequencies, which were significantly negatively correlated with intact but not total HIV DNA. For percentage of intact HIV DNA, negative correlations were observed with the frequencies of 2 subsets of bifunctional HIV peptide-stimulated CD4 T cells (IL-2+TNFα+ and IFNγ+IL-2+), as well as years on ART. Changes in immune markers in this study relative to years of ART were also indicative of progressive immune recovery (Figure S2-S4 and Table S4). Specifically, years of ART negatively correlated with CD38 expression on CD8 T cells and positively correlated with the frequency of CD8 Tcm cells as well as CD107+IL-2+IFNγ-TNFα-HIV-specific CD8 T cells (Table S4). There are likely many more immune parameters that change in concert with ART as recently reported (47); however, the noise associated with cross-sectional immunophenotype data likely precludes their identification in this study.

Leave-One-Covariate-Out (LOCO) inference identifies specific HIV reservoir and clinical-demographic parameters important for the prediction of several immunophenotypes

Given the evidence that reservoir and immune recovery dynamics likely occur in concert, we next used a variable importance approach - Leave-One-Covariate-Out (LOCO) inference analysis, to account for potential confounding variables and more carefully assess the relative importance of a given variable to an immunophenotype. The LOCO inference approach is demonstrated in Figure 2A. First, a least-squares regression linear model that predicts the dependent variable is fitted on a set of independent variables, and an R2 value is generated (Figure 2A Step 1). Next, one independent variable is excluded from the model, the model is refitted, and an adjusted R2 value is determined (Figure 2A Step 2). Finally, the change in R2 scores, ΔR2, is calculated (Figure 2A Step 3). This process is then repeated for all independent variables in the model, and the ΔR2 values are plotted in a heatmap for the variables of interest (Figure 2B-D).

Leave-one-covariate-out (LOCO) analysis for clinical-demographic features and reservoir characteristics while predicting immunophenotypes.

A. Explanation of leave-one-covariate-out (LOCO) analysis based on example of %Tn CD8 T for clinical-demographic features and reservoir characteristics while predicting immunophenotypes. Analysis was performed for all 133 immunophenotypes considered in the study. The top 10 biggest drops in adjusted R2 scores are reported for models that use total reservoir size (B.), intact reservoir size (C.), or percent intact (D.) as features in addition to clinical and demographic information. Participants with missing years of ART values are excluded from this analysis. The missing value of the CD4 nadir for one participant is imputed. In Figure S6, the LOCO analysis results are visualized for all 133 immunophenotypes.

We applied the LOCO inference analysis to all 133 immunophenotype parameters (dependent variables). Three separate analyses were conducted for a set of independent variables consisting of age, sex, years of ART, CD4 nadir, recent CD4 account, years of HIV infection prior to ART (=NA, < 1, ≥ 1), and 1 of 3 HIV reservoir metrics (total HIV DNA, intact HIV DNA, or percentage intact HIV DNA).

Multicollinearity analysis confirmed that independent variables from this analysis were not correlated to a degree that would interfere with model generation (Table S5). Each analysis produced a heatmap of 133 dependent immunophenotype variables, with visualization of the variables for each immunophenotype (Figure S6, Tables S6-S8). For each of these analyses, we displayed the top 10 variables with the largest ΔR2 value for LOCO inference incorporating either total (Figure 2B), intact (Figure 2C), or percentage intact (Figure 2D) HIV DNA. In general, the least-squares linear regression models were at best weakly explanatory (R2 values in range [-8.38%, 31.69%]) of the global variability in immunophenotypes (Tables S6-S8, Figure 2B-D, Figure S7), however, the LOCO analysis is still valuable for understanding which clinical/demographic variables and HIV characteristics are important to the immunophenotypes.

Across all 3 analyses, age emerged as an important covariate of several immunophenotypes related to T cell subsets, particularly the naïve and central memory compartments, as well as KLRG-1, a marker of immune senescence (Figure 2 and Figure S6). This is consistent with well-described age-related changes in T cell subset populations (49, 50). Notably, biological sex was a key variable for model prediction of NK cell frequency and NK cell NKG2A expression. Years of ART was important for predicting the frequencies of HLA-DR+ CD8 T cell subsets including Tcm, Tn, and Tem, but more so for the model incorporating intact, rather than total, HIV DNA (Figure 2B and 2C). This suggests that both years on ART and total HIV DNA are important explanatory variables for the variance in CD8 T cell activation levels. Total HIV DNA was also noted to be an important predictor of PD-1+ CD4 T naïve cells, HIV-specific IL-2+TNFα-IFNγ-CD107-CD4 T cells, and CD4/CD8 T cell frequencies, among others (Figure 2B, Figure S6). Total HIV DNA and, to a lesser extent, intact HIV DNA were important predictors of CD4 T effector memory HLA-DR+CD38+ expression (Figure 2B and 2C). In general, total and intact HIV DNA were identified to be predictive of similar immune features (Figure S6, Tables S6-S8); how-ever, total HIV DNA generally had greater magnitude ΔR2 than intact HIV DNA. The percentage of intact HIV DNA did not appear to contribute greatly to the predictive power of the model for the variance in the immunophenotypes examined (Figure S6C, Tables S10, S11), most likely due to its relatively higher correlation with years of ART (Figures 1C). Nonetheless, the LOCO inference findings suggest a complex biological concert of immunologic changes that occur over years of life and suppressive ART, and which may be influenced by the HIV reservoir.

Receiver operator characteristic curve analysis demonstrates distinct immune parameters associated with intact and total reservoir frequencies

The complex relationship between the HIV reservoir and the immune system led us to hypothesize that machine learning (ML) approaches that combine several parameters into models of the HIV reservoir could be a useful approach to understand this interaction. Before this could be attempted, we decided to first identify a defined set of the most valuable variables from which ML models could be built. To achieve this, we first binarized the 3 reservoir metrics (total, intact, %intact) into high (above median) or low (below median) reservoir groups and generated receiver operator (ROC) curves for the 144 clinical-demographic and immunophenotype parameters (Figure 3). The area under the curve (AUC) of the ROC curve indicates the ability of the feature (immune parameter or clinical/demographic information) to correctly identifying a participant as having qualitatively low or high total HIV DNA (Figure 3A), intact HIV DNA (Figure 3B), or percentage intact HIV DNA (Figure 3C). A model that randomly guesses high or low reservoir has an AUC of 0.5. In this analysis, 57 host variables had an AUC higher than 0.6 for one or more of total HIV DNA, intact HIV DNA, or percent intact DNA (Table S9). The most effective individual immune markers for classification of high versus low total HIV DNA included %NKG2A+ CD4 T (AUC = 0.70), %PD-1+ Tn CD4 T (AUC = 0.68), and %CD38+/HLA-DR-CD8 T (AUC = 0.68). In contrast, the most effective markers for classifying based on the frequency of intact HIV DNA included %CD127+ CD4 T (AUC = 0.71), %CD8 T (AUC = 0.66), and %Tn CD8 T (AUC = 0.65). Finally, for percentage intact HIV DNA, years of ART (AUC = 0.72), %CD107-IFNγ-IL-2+TNFα+ CD4 T (AUC = 0.65) and %KLRG1+/PD1+ CD4 T (AUC = 0.65) were the most effective (Figure 3).

Receiver operating characteristic (ROC) curves identify PWH parameters that can classify reservoir characteristics.

For total reservoir (A), intact reservoir (B), and percent intact (C), ROC curves are plotted for all 144 immune markers, demographics, and clinical variables (shown in gray). Axes represent the true positive rate (TPR) and the false positive rate (FPR) for each variable for classifying study participants into low (below median) versus high (above median) reservoir size. ROC curves for ten variables with the highest AUC values are shown in color for every HIV reservoir characteristic. Striped black lines represent the ROC curves of a random model. For years of ART ROC curves, we exclude participants with missing years of ART values.27

Similar to the findings in the Spearman correlations (Table 2) and LOCO inference analysis (Figure 2), there were more variables that had AUC values > 0.6 for total reservoir size (Table S9): 44 variables above this threshold for the total reservoir, but only 19 variables for the intact reservoir and 23 variables for percent intact. Overall, this approach allowed us to derive a ranked list of the most predictive immune parameters for each aspect of the HIV reservoir, and these highly ranked features were thus used for subsequent dimension reduction and ML modeling.

Dimension reduction reveals two clusters of PWH with distinct HIV reservoirs

To examine the overall structure of dataset we employed an unsupervised dimension reduction (DR) machine learning approach (PaCMAP (51)) using the 10 immune cell features with the highest AUC values for classifying participants based on total, intact, or percentage intact HIV DNA (Figure 3, Table S9). Interestingly, while no clear clustering was observed when we use intact or percentage intact HIV DNA associated features, we observed two distinct clusters of PWH (cluster 1 and cluster 2, Figure 4A) when using total HIV DNA associated features. These clusters were of roughly equal size with 50 participants in cluster 1 and 65 in cluster 2. Projecting total reservoir size (above or below the median) onto the clustering plot, we observe a strong distinction between the two clusters in terms of total HIV DNA reservoir characteristics, (Figure 4B,C). Analysis of quantitative reservoir frequencies across clusters (Figure 4C-E) demonstrated that cluster 1 is characterized by a smaller absolute intact reservoir frequency, but greater percentage of intact proviruses (Figure 4C-E). In contrast, cluster 2 is defined by a larger absolute intact reservoir frequency, but lower percentage of intact proviruses (Figure 4C-E).

Dimension reduction reveals two major clusters of PWH with distinct immune systems and reservoirs.

A. PaCMAP was applied to the data using the ten immune cell features with the highest AUC values for classifying participants based on total reservoir size, and two clusters (cluster 1 and cluster 2) are identified. B. Same as A but data points are color-coded by total reservoir size (high=purple, low=pink). Total reservoir size (C.), intact reservoir size (D.), and percent intact (E.) is shown for participants within each cluster. F. Key immune cell features that distinguish cluster 1 from cluster 2 are identified by visualizing the features with the highest AUC values with respect to classifying cohort participants based on cluster membership. Axes represent the true positive rate (TPR) and the false positive rate (FPR) for each variable. Immune markers and clinical-demographic features are shown for each cluster in Figure S8.

To gain insight into the immune cell features that distinguish these two clusters, we generated ROC curves for the 133 immune parameters based on their ability to identify membership in cluster 1 versus cluster 2 for each participant (Figure 4F). Many features that distinguished the clusters overlapped with the features selected for the clustering analysis, including expression patterns of CD38, HLA-DR, KLRG-1 and PD-1 on CD4 and CD8 T cells and CD4 T naïve cell frequencies. However, some novel features emerged, including CD4 KLRG1-CD27+ frequency, CD8 KLRG1-PD1-frequency, and CD8 Tn frequency (Figure 4F and Figure S8). Cluster 2, which had a higher total and intact absolute reservoir size but lower percentage of intact proviruses, was associated with a lower fraction of naïve CD4 and CD8 T cells. Additionally, cluster 2 had generally higher expression of immune exhaustion markers, including KLRG1 and PD-1 (Figure S8). These data indicate that cluster membership is highly associated with the overall level of immune activation and exhaustion for the individual. We have previous shown that cannabis use reduces the overall level of immune activation and inflammation in PWH, despite having minimal impact on the size of the HOV reservoir (52). Interestingly, a subset of participants in this study had clinical data regarding cannabis (CB) use. CB users were enriched in cluster 1, which had lower levels of immune exhaustion markers (Figure S8). Notably, with regard to other clinical variables, cluster 2 had an older median age, longer time on ART, lower CD4 nadir, and lower current CD4 T cell count (Figure S8).

Overall, these data demonstrate that, within the cohort of PWH on long-term ART, there were two major clusters of participants with distinct reservoir characteristics and immunophenotypes. These two clusters may reflect the results of a complex concert of age-related immune changes, immune recovery following years of ART, and decay of the intact HIV reservoir, among other possibilities.

Decision tree visualization of PWH with respect to reservoir characteristics

Since the interaction of the immune system and the HIV reservoir is multifactorial, we hypothesized that models that consider multiple parameters simultaneously could more accurately describe the overall dataset, and provide insights regarding the biology of the HIV reservoir and the host immune system. To accomplish this, we employed a decision tree approach to visualize combinations of variables that classify participants as having high (above median) or low (below median) reservoir size. As compared to the dimension reduction technique, decision tree visualization is an interpretable supervised approach and does not require post-hoc analysis.

We first selected 35 variables with the highest ROC AUC values for either total, intact, or percentage intact HIV DNA to be considered for model generation (Figure 4). Using these parameters, we fitted Generalized and Scalable Optimal Sparse Decision Trees (GOSDT) (53) to the data in order to classify high versus low total HIV DNA, intact HIV DNA, or percentage intact HIV DNA. We required the trees to achieve at least 80% accuracy for classifying PWH in the cohort, as well as have at least five PWH in each leaf. Since these trees are based on the entire dataset, these models are thus descriptive rather than predictive.

For the total HIV DNA decision tree, only four immune variables were required to accurately describe high versus low HIV DNA status (Figure 5A, Figure S9): CD8 T cell frequency, CD4 nadir, %CD38+HLA-DR-CD8 T cells, and %NKG2A+ CD4 T cells. The tree divided the cohort into five sub-groups (leaves), among which three have high total reservoir size and two have low total reservoir size. Comparing the labels provided by GOSDT model with the actual data, the tree achieved 83.5% accuracy (i.e. misclassifying 19 PWH among the overall cohort of 115 PWH). Notably, when we combined all samples from “high total reservoir” leaves and all samples from “low total reservoir” leaves, we observed a significant difference in the actual median total reservoir size for these two groups (266/M for low total and 1288.5/M for high total, Mann Whitney U test p-value is 3.56e-13, Figure 5B).

Decision tree visualization of the association of immune cell subsets with reservoir characteristics.

(A, C, E) Host variables (immune cell frequencies, demographic, and clinical information) were used to visualize the PWH dataset using the optimal sparse decision trees algorithm GOSDT. The overall set of PWH was classified as likely having high (above median, orange “leaves”) or low (below median, blue “leaves”) total reservoir size (A), intact reservoir size (C), and percent intact (E). In each leaf, “med” denotes the median HIV characteristic of PWH, N is the number of PWH in the leaf, and MN is the number of misslabeled PWH. (B, D, F) PWH in model leaves associated with high (orange) or low (blue) reservoir size characteristics were aggregated and a Mann-Whitney U test was performed to determine statistical significance between the actual total reservoir size of the “high” and “low” groups for total reservoir size (B), intact reservoir size (D), and percent intact (F). For the percent intact tree we exclude participants with missing values of years of ART. For total and intact reservoir size, missing values of years of ART were imputed, however, since the trees do not use this variable, imputations do not influence results. Visualization trees are explained with sets of rules in Supplemental Materials: Figure S9 for total reservoir size, Figure S10 for intact reservoir size, and Figure S11 for percent intact.

We then repeated this approach for classifying the cohort participants with respect to intact reservoir size (Figure 5C, Figure S10). Despite the overall lower correlation between the immune cell phenotypes and the intact reservoir size, this tree nevertheless achieved 82.6% accuracy. The tree had six leaves (three with low intact reservoir, three with high intact reservoir) and relied on %Tn CD8 T cells, %CD107-IFNγ+IL-2+TNFα-CD4 T cells, %CD127+ CD4 T cells, %CD38-HLA-DR+ Tcm CD8 T cells, and CD4 Nadir. When we combined the high intact leaves together and the low intact leaves together, we observed a significant difference in median intact reservoir frequency between the groups (29.5/M for low intact and 101/M for high intact, Mann Whitney U test p-value is 9.59e-08, Figure 5D).

Finally, a third iteration of GOSDT generation was performed, classifying the cohort participants with respect to percentage intact of total proviruses (Figure 5E, Figure S11). This tree achieved 82.4% accuracy and relied on years of ART treatment, %KLRG1+CD27-CD4 T cells, %CD127+ CD4 T cells, and recent clinical CD4 count. This tree has two leaves with low percent intact reservoir and three leaves with high percent intact reservoir. We observed a significant difference in the actual percent intact values between the model generated high percent intact and low percent intact leaves (median 5.7% for low percent intact leaves and median 15.3% for high percent intact leaves: Mann Whitney U test, p-value is 5.5e-10, Figure 5F).

Collectively, these decision trees highlight combinations of immune parameters that can accurately describe qualitatively high versus low total, intact, and percentage intact HIV DNA in a cohort of n = 115 PWH. This visualization serves as a basis for mechanistic hypotheses about the interactions of the immune system and HIV reservoir during long-term ART.

Machine learning algorithms identify immune feature combinations that predict high versus low reservoir metrics with ∼70% test accuracy

Although clustering and decision tree analysis permit visualization and understanding of global structures within a dataset, we were curious if combinations of immune and clinical-demographic parameters could actually accurately predict (rather than only visualize) whether a given participant had a high (above median) or low (below median) values with respect to total HIV DNA, intact HIV DNA, or percent intact HIV DNA. We considered five machine learning algorithms including Logistic Regression with L2 regularization (LR), CART, Support Vector Machines with RBF kernel (SVM), Random Forest (RF), and Gradient Boosted Trees (GBT). For reservoir size characteristics, we measured accuracy of the models over 10 random splits of the data into training and test sets (Figure 6A and Table S10). Overall, we found that for total reservoir size Logistic Regression achieved highest mean classification accuracy (69.31%) in test data. The accuracy of these models is likely limited by the sample size (n=115) and noise in the data, as evidenced by lower R2 score of machine learning models for non-binarized HIV characteristics (Figure S12). To examine the contribution of individual immune features to model performance, we examined logistic regression coefficients for each immune cell variable in the model. Since Logistic Regression coefficients are associated with expected change in log odds (based on loge), we can think about the coefficient β for variable X in the following way: increasing variable X by one unit multiplies the odds of high reservoir size (probability that the reservoir size is high divided by the probability that the reservoir size is low) by eβ.

Predicting HIV reservoir characteristics with machine learning.

Average training and test accuracies over ten training and test data splits for Random Forest (RF), Gradient Boosted Trees (GBT), Support Vector Machines with RBF kernel (SVM), Logistic Regression (LR), and CART models for total reservoir size (A), intact reservoir size (C), and percent intact (E) are shown. For one split of training and test sets, logistic regression models are visualized for total reservoir (B), intact reservoir (D), and percent intact (F). On the y-axis we show variables used by the model, while the x-axis displays coefficient values for individual variables used by models. For percent intact models we exclude participants with missing values of years of ART. For total and intact reservoir size, missing values of years of ART were imputed. The missing value of the CD4 nadir for one participant was imputed as well using the MICE algorithm.

In Figure 6B we visualize the logistic regression model for one data split among ten we considered for total reservoir size. For this split we observe that higher values of %NKG2A+ CD4 T, %PD-1+ Tn CD4 T, and %Tcm CD8 T are associated with a increased probability of total reservoir size being high. On the other hand, an increase in %Tn CD4 T decreases the odds of high reservoir size. The model visualized in Figure 6B achieved 75.86% training and 75% test accuracy.

Similarly, logistic regression models performed well for predicting intact reservoir size compared to other methods (average 65.17% test accuracy, see Figure 6C). In Figure 6D we visualize the Logistic Regression model for one fixed data split. In this model we observed that higher values of %CD107-IFNγ+IL-2+TNFα-CD4 T and %CD127+ CD4 T, and % CD4 T are associated with lower probability of intact reservoir size being high, while higher values of %KLRG1+CD27+ CD8 T are associated with increased probability of intact reservoir size being high. For analysis of percentage intact HIV DNA, we display training and test accuracy values in Figure 6E. For the visualized model for one data split (Figure 6F), an increase in %CD127+ CD4 T, years of ART, %KLRG1+PD-1+ CD4 T, and %Tem CD4 T leads to a lower probability of percent intact HIV DNA being high.

Overall, these analyses demonstrate that we can use machine learning tools to construct models that can predict with approximately 70 percent accuracy whether a given PWH has qualitatively low or high total, intact or percentage of HIV DNA. The coefficients identified in each of these models, including NKG2A and CD127 expression on CD4 T cells, identify specific immunologic nodes that serve as a basis for the generation of hypotheses about the interplay between total versus intact HIV DNA and the host the immune system.

Discussion

In this cross-sectional study of peripheral blood samples from a well-characterized clinical cohort of 115 PWH on long-term ART, we used IPDA and high parameter flow cytometry to better understand the associations between the HIV reservoir and a broad range of immunophenotypes related to T cell differentiation, activation, exhaustion, senescence, HIV-specific T cell responses, as well as NK cell phenotypes. Using data science approaches, specific HIV reservoir features (total, intact, percentage intact HIV DNA) and clinical-demographic variables (age, sex, CD4 nadir, recent CD4 count, years on ART, years of HIV infection prior to initiating ART) were identified that were particularly important for predicting specific immunophenotypes in PWH on long-term ART. Distinct host immune correlates of intact versus total HIV DNA were also identified. Further, machine learning approaches enabled identification of combinations of immune parameters that accurately classified and predicted whether a given PWH is likely to have qualitatively high or low total, intact, or percentage intact HIV DNA. Collectively, this study provides novel insights into HIV reservoir biology and notably also demonstrates the utility of specific data science and machine learning approaches for HIV reservoir studies. Development and application of these analytic approaches are important to enable biological insights from high-parameter studies of the HIV reservoir and other complex biology, particularly with the advent of high-dimensional techniques such as CYTOF and scRNAseq.

Previous studies have sought to identify immune and other correlates of the total HIV DNA reservoir (26–45). These studies have identified several activation/exhaustion markers that correlate with HIV reservoir frequency, including T cell expression of HLA-DR, CD38, PD-1, CTLA-4, and LAG-3, the CD4/CD8 ratio, and NK cell KLRG1 expression. However, not all studies have found associations of total HIV DNA and immune subsets, particularly for CD4 T cell HLA-DR and CD38 expression(40, 54–56) during long-term ART. The correlative results from this present study corroborate many of these studies, and provide additional insights. A number of correlations between total HIV DNA and immune activation and exhaustion were identified. There were weak to moderate positive correlations observed between total HIV DNA and CD4 T cell expression of NKG2A, PD-1, KLRG1, CD27, HLA-DR as well as CD8 T cell expression of PD-1, CCR7, and CD38. Total HIV DNA was inversely correlated with CD8 T cell expression of CD38, CD4 T cell expression of CD27, the frequency of naïve CD4 and CD8 T cells, and the CD4/CD8 ratio.

Interestingly, when we examined immune correlates of intact HIV DNA measured by the IPDA, many of the same relationships as described above were noted, but with decreased magnitude and lacking statistical significance, with the exception of the inverse correlation observed for CD4/CD8 T cell ratio and intact HIV DNA. This is consistent with a recent study using the IPDA that did not find associations of intact HIV DNA with CD4 T cell HLA-DR and CD38 co-expression (57), as well as a study using nearfull-length sequencing of single cells that found only slight enrichment of HLA-DR or PD-1 in cells with genetically intact, inducible proviruses (37).

Uniquely, however, CD127 expression on CD4 T cells was significantly inversely associated with intact reservoir frequency. CD127 is the alpha-chain IL-7 receptor and is expressed on long-lived memory T cells that are depleted during untreated HIV infection. Interestingly CD127+ cells have recently been found to be highly susceptible to latent HIV infection in tissues (58), and IL-7 is associated with slower natural reservoir decay (27). Furthermore, IL-7 signaling is known to drive expression of VLA-4, which was recently reported to highly enrich for genetically intact, inducible proviruses (37). This observational finding that intact but not total HIV DNA inversely correlates with CD127 expression on CD4 T cells requires further investigation. Finally, this initial immune correlate analysis demonstrated that the percentage of intact HIV DNA was negatively correlated with years of ART as well as and HIV-specific CD4 T cell responses. These findings are consistent with known selective decay of intact proviruses over years of ART therapy (57, 59, 60).

Dimension reduction machine learning approaches identified two robust clusters of PWH when using total HIV DNA reservoir-associated immune cell frequencies (Figure 4A), but not for intact or percentage intact HIV DNA (Figure 4B and 4C). This pattern is likely indicative of a more profound association of total HIV DNA with host immunophenotype relative to intact HIV DNA. This is certainly plausible giving the growing appreciation for the immunogenicity and high frequency of defective proviruses within the HIV reservoir (61–64). These clusters exhibited distinct reservoir characteristics, with one cluster being enriched with participants with a larger total reservoir size, a lower percentage of intact viruses, a lower frequency of naïve T cells and elevated expression of activation and exhaustion markers, while the other contained PWH with smaller total reservoirs, higher percentage of intact viruses, higher frequency of naïve T cells and lower expression of activation and exhaustion markers. The basis for the two distinct clusters is unclear, but the cluster membership was notably associated with age and time on therapy. The existence of two distinct clusters of PWH with different immune features and reservoir characteristics could have important implications for HIV cure strategies – these two groups may respond differently to a given approach, and cluster membership may need to be considered to optimize a given strategy. It is tempting to speculate that these two clusters represent differing severity of the initial immune insult of untreated HIV infection, and possibly dichotomous degrees of immune recovery following initiation of ART. These two clusters are also interesting in the context of a previous study examining reservoir characteristics which proposed two major types of viral reservoir within PWH. One type with smaller reservoirs that were enriched in Tcm cells and associated with lower levels of Ki67+ cells and immune activation, and another type with larger reservoirs that were enriched in Ttm cells and associated with higher levels of Ki67+ cells and immune activation (27). It is possible that these two proposed types of reservoirs represent the two clusters we observe using immune markers and dimension reduction, although we were unable to examine the frequency of HIV proviruses specifically within Tcm and Ttm cells in this study.

In addition to unsupervised dimension reduction approaches, we also found that we were able to construct simple, interpretable decision trees that could describe our cohort of PWH with more than 80% accuracy with respect to the size of the reservoir. Furthermore, when we constructed decision trees to classify PWH separately based on total reservoir, intact reservoir and percentage intact, distinct sets of features were important (though some were the same) for each of these models. The decision tree analyses highlight complex immune parameter combinations that serve as a basis for unique hypotheses about HIV reservoir biology. For instance, lower levels of HIV-specific CD4 T cell responses and lower levels of CD8 Tcm HLA-DR expression were useful for identifying individuals with higher frequencies of intact HIV DNA. This suggests that individuals with higher levels of intact proviruses may have lower levels of HIV-specific immune responses and memory CD8 T cell activation that could facilitate clearance of intact proviruses. Notably, the same relationships were not observed for the total HIV DNA decision tree, highlighting the likely more potent immunogenicity of intact proviruses.Another striking example of machine learning approaches identifying potentially insightful reservoir biology is the percentage intact HIV DNA decision tree. As expected, years of ART was critical to predict the percentage of intact proviruses, but CD4 T cell CD127 expression, CD4 T cell KLRG expression and recent CD4 count also played a role. Notably, following approximately 7 years of ART, lower CD4 T cell CD127 expression or low recent CD4 T cell count were useful to classify individuals with higher percentages of intact proviruses. We hypothesize that this finding suggests incomplete immune recovery (low CD4 T cell CD127 expression and low CD4 T cell count) is associated with a larger percentage of intact proviruses, perhaps because of a lack of dilution with de novo naïve CD4 T cells over many years of ART.

Logistic regression models that we trained to classify high versus low total reservoir, intact reservoir, and percentage intact were able to predict out-of-sample with ∼70% accuracy on average. Each of these models relied on five features only and were simple enough for us to visualize and assess how important each variable is for every characteristic of reservoir size. These models corroborated the potential biological relationships identified in the decision tree analysis, and confirmed that these variables could actually be used to predict out-of-sample whether total, intact or percentage intact HIV DNA was qualitatively high or low.

Our findings should be considered in the light of some inherent limitations and caveats. This study is cross-sectional in nature and is primarily observational, so caution should be used interpreting findings associated with time on therapy. As with any observational study, confounding bias may influence our results, particularly for correlative analyses. Another limitation to this study is that our analyses are limited to peripheral blood, and may not represent HIV reservoirs and immune cells located in tissues known to harbor latent HIV such as gut, spleen and brain.

Nevertheless, our study does have some strengths compared to previous work in this area, specifically a highly detailed flow panel, allowing us to quantify numerous immune subsets with a high degree of resolution. Also, the use of the intact proviral DNA assay allows us to separately examine the association of the immune system with intact and total viral DNA. It is however, important to note that IPDA does not measure all proviral sequences in PWH and is likely failing to identify some sequences due to proviral polymorphisms or dual deletions in the two IPDA amplicons. Also, a minor fraction of IPDA+ proviruses are actually defective due to the limited specificity of using two short amplicons to select for full-length intact proviruses (48, 65).

Overall, these findings suggest a complex concert of immune recovery, HIV reservoir dynamics, and intrinsic host factors (age, biological sex) that shapes host immunophenotype, even after years of ART. Mechanistic work will be needed to fully dissect the dynamic relationship between the immune system and the reservoir. Nevertheless, these findings support a model in which ongoing interaction between the HIV reservoir and the host immune cells continue to drive an association, albeit a relatively minor one, with persistent CD4 T cell activation during long-term ART. Notably, both intact and defective proviruses appear to contribute to this immune signature of HIV latency during long-term ART.

This study identifies several areas for future investigation. The question of whether the HIV reservoir directly drives persistent immune activation could be addressed directly by specifically suppressing viral transcription during therapy and examining the impact on the immune system. Future clinical trials with effective latency reversal and clearance regimens could include measurement of HIV-specific T cell responses, as this analysis suggests that lower levels may be associated with reduced intact HIV DNA clearance.

There is also growing appreciation for additional factors that correlate with the HIV reservoir. Specifically, the size of the reservoir has been linked to the abundance of specific bacteria species in the intestinal microbiome (66). Also, genetic studies have indicated the existence of single nucleotide polymorphisms (SNPs) that are associated with reservoir size, including SNPs in MX1 and IRF7 (67–69). These, along with future studies exploring the interplay of the host immune system and HIV reservoir during long-term might leverage new high-dimensional technologies such as scRNAseq and mass cytometry. The machine learning approaches described and applied in this study may be particularly useful for gaining biological insights into these high parameter datasets, particularly using analysis of global patterns in the data that may be otherwise imperceptible to humans.

Methods

Cohort and sample collection

115 people with HIV (PWH) were recruited from two clinical sites. 66 PWH were recruited at Duke University medical center and 49 PWH at UNC Chapel Hill. Peripheral blood was collected and peripheral blood mononuclear cells (PBMCs) were obtained by Ficoll separation, then frozen in 90% fetal bovine serum 10% dimethylsulfoxide (DMSO) (52, 59, 70, 71).

Flow cytometry

Cell preparation and staining followed previously described methods (72). All antibodies were titrated to optimize signal-to-noise ratio on PBMCs prior to use, assuming a 50μL staining volume. Cell viability was examined using Zombie-NIR Fixable Viability Dye (0.4μL per 50μL staining volume; Biolegend). Antibodies used for surface staining were as follows: KLRG1-BV421 (SA231A2 clone, Biolegend, CD45RA-PacBlue (H100 clone, BL), CD8-BV570 (RPA-T8 clone, Biolegend), CD127-BV605 (A019D5 clone, Biolegend), CD56-BV650 (5.1H11 clone, Biolegend), CCR7-BV711 (G043H7 clone, Biolegend), CD27-BV750 (O323 clone, Biolegend), PD1-VioBright515 (REA165 clone, Miltenyi), NKG2A-PE-Vio615 (REA110 clone, Miltenyi), CD16-PerCP-Cy5.5 (33G8 clone, Biolegend), CD38-PCPeF710HB7 clone, TF), CD14-SparkNIR685 (63D3 clone, Biolegend), CD19-SparkNIR685 (HIB19 clone, Biolegend), HLA-DR-APC-F750 (L243 clone, Biolegend). Antibodies used for intracellular staining were as follows: CD3-BV480 (UCHT1 clone, BD), CD4-PerCP (L200 clone, BD), IFN-g-PE-Cy7 (4S.B3 clone, Biolegend), IL-2-APC (MQ1-17H12 clone, Biolegend), TNFα-AF700 (Mab11 clone, Biolegend). Samples were analyzed using a Cytek Aurora spectral flow cytometer. Flow cytometry analysis was performed in FlowJo v10.8 software.

Intact Proviral DNA Assay (IPDA)

Cryopreserved samples of PBMCs from each study participant were viably thawed. A portion was used for immunophenotyping as described above, and the remainder were subjected to total CD4 T cell negative selection with the StemCell Technologies EasySepTM Human CD4+ T Cell Enrichment Kit (Cat# 19052). CD4 T cell DNA was extracted using the QIAamp DNA Mini Kit and quantified on a NanoDrop 1000 (Thermo Fisher Scientific). IPDA was performed as originally described (48), with a validated PCR annealing temperature modification to increase signal to noise ratio (59). Gating for positive droplets was set using negative (DNA elution buffer and HIV-seronegative CD4 T-cell DNA), and positive (Integrated DNA Technologies gblock amplicon) positive control wells processed in parallel. DNA shearing index values were similar to those reported previously (median, 0.33; interquartile range [IQR], 0.31–0.34) (59, 70, 71). A median of 1.04 x 106 (Q1 8.71 x 105, Q3 1.16 x x 106) cell equivalents were assessed for each donor.

Statistics

Variable data analysis

To create binarized labels that represent reservoir characteristics, we split the data at the median, which is 553/M for total reservoir size, 53/M for intact reservoir size, and 8.64 percent for percent intact. For every variable (immune cell frequency, demographics, and clinical information), we computed Spearman correlation and an Area Under the Curve (AUC) value. To compute the AUC value, we first created a Receiver Operating Characteristic (ROC) curve by plotting the true positive rate versus the true negative rate for every cell subset frequency. AUC is then computed as the area under the ROC curve using the trapezoidal rule.

To assess the importance of clinical, demographic information and HIV reservoir characteristics in describing the immune markers, we performed a leave-one-covariate-out (LOCO) analysis. We trained a linear regression model (M1) with intercept based on all variables. We removed each variable at a time and trained a new model without this variable. Then we computed the difference in R2 score before and after dropping the variable. Variance Inflation Factors (VIFs) for age, sex, years of ART, CD4 nadir, recent CD4 account, and years of HIV infection prior to HIV infection were less than 2, indicating an acceptable level of correlation between these independent variables (73). However, higher VIF values were observed for intact and total HIV DNA, therefore separate LOCO inference analyses were conducted in order to avoid artificial fluctuations in model fit due to multicollinearity.

Data visualization

We used PaCMAP (51) to reduce the dimensionality of the dataset to a 2-dimensional space. Due to the relatively small sample size of our dataset (n=115), we set the number of neighbors to 5. Additionally, we used GOSDT (53) which computes sparse optimal trees to identify patterns in the dataset. For all the visualization trees we set the depth budget parameter to 4 and the regularization parameter to 0.02.

Training and generalization

Our training procedure for classification is described in Algorithm 1 in Supplemental Materials, which is a standard machine learning pipeline with an added search for the number of variables. For the full hyperparameters list see Table S8. We normalized continuous variables to have values between 0 and 1. For logistics regression models, we set the regularization parameter of .e − 2 regularization based on cross-validation. The values are 10 for total reservoir size and percent intact and 100 for intact reservoir size.

Study approval

Written informed consent was obtained for all study participants. The study design was reviewed and approved by IRB for both Duke University and UNC Chapel Hill.

Data availability

Deidentified datasets will be made publically available upon manuscript publication.

Author contributions

Designed the study – EPB, DMM, CDR. Conducted experiments – SDF, ADVC. Analyzed the data – LS, YW, SDF. Wrote the manuscript: All authors.

Conflict of interest statement

The authors have declared that no conflict of interest exists.

eLife Assessment

This useful paper looks for correlations between immunophenotypic markers and several measures of HIV reservoir volume in cross-sectional cohorts of people living with HIV on ART using several bioinformatic and machine-learning tools. The level of evidence linking these variables is incomplete given possible confounding variables, lack of directionality & effect size, and mechanistic basis.

Reviewer #1 (Public Review):

Summary:

Semenova et al. have studied a large cross-sectional cohort of people living with HIV on suppressive ART, N=115, and performed high dimensional flow cytometry to then search for associations between immunological and clinical parameters and intact/total HIV DNA levels.

A number of interesting data science/ML approaches were explored on the data and the project seems a serious undertaking. However, like many other studies that have looked for these kinds of associations, there was not a very strong signal. Of course, the goal of unsupervised learning is to find new hypotheses that aren't obvious to human eyes, but I felt in that context, there were (1) results slightly oversold, (2) some questions about methodology in terms mostly of reservoir levels, and (3) results were not sufficiently translated back into meaning in terms of clinical outcomes.

Strengths:

The study is evidently a large and impressive undertaking and combines many cutting-edge statistical techniques with a comprehensive experimental cohort of people living with HIV, notably inclusive of populations underrepresented in HIV science. A number of intriguing hypotheses are put forward that could be explored further. Sharing the data could create a useful repository for more specific analyses.

Weaknesses:

Despite the detailed experiments and methods, there was not a very strong signal for the variable(s) predicting HIV reservoir size. The Spearman coefficients are ~0.3, (somewhat weak, and acknowledged as such) and predictive models reach 70-80% prediction levels, though sometimes categorical variables are challenging to interpret.

There are some questions about methodology, as well as some conclusions that are not completely supported by results, or at minimum not sufficiently contextualized in terms of clinical significance.

On associations: the false discovery rate correction was set at 5%, but data appear underdetermined with fewer observations than variables (144vars > 115ppts), and it isn't always clear if/when variables are related (e.g inverses of one another, for instance, %CD4 and %CD8).

The modeling of reservoir size was unusual, typically intact and defective HIV DNA are analyzed on a log10 scale (both for decays and predicting rebound). Also sometimes in this analysis levels are normalized (presumably to max/min?, e.g. S5), and given the large within-host variation of level we see in other works, it is not trivial to predict any downstream impact of normalization across population vs within-person.

Also, the qualitative characterization of low/high reservoir is not standard and naturally will split by early/later ART if done as above/below median. Given the continuous nature of these data, it seems throughout that predicting above/below median is a little hard to translate into clinical meaning.

Lastly, the work is comprehensive and appears solid, but the code was not shared to see how calculations were performed.

Reviewer #2 (Public Review):

Summary:

Semenova et. al., performed a cross-sectional analysis of host immunophenotypes (using flow cytometry) and the peripheral CD4+ T cell HIV reservoir size (using the Intact Proviral DNA Assay, IPDA) from 115 people with HIV (PWH) on ART. The study mostly highlights the machine learning methods applied to these host and viral reservoir datasets but fails to interpret these complex analyses into (clinically, biologically) interpretable findings. For these reasons, the direct translational take-home message from this work is lost amidst a large list of findings (shown as clusters of associated markers) and sentences such as "this study highlights the utility of machine learning approaches to identify otherwise imperceptible global patterns" - lead to overinterpretation of their data.

Strengths:

Measurement of host immunophenotyping measures (multiparameter flow cytometry) and peripheral HIV reservoir size (IPDA) from 115 PWH on ART.

Major Weaknesses:

(1) Overall, there is little to no interpretability of their machine learning analyses; findings appear as a "laundry list" of parameters with no interpretation of the estimated effect size and directionality of the observed associations. For example, Figure 2 might actually give an interpretation of each X increase in immunophenotyping parameter, we saw a Y increase/decrease in HIV reservoir measure.

(2) The correlations all appear to be relatively weak, with most Spearman R in the 0.30 range or so.

(3) The Discussion needs further work to help guide the reader. The sentence: "The correlative results from this present study corroborate many of these studies, and provide additional insights" is broad. The authors should spend some time here to clearly describe the prior literature (e.g., describe the strength and direction of the association observed in prior work linking PD-1 and HIV reservoir size, as well as specify which type of HIV reservoir measures were analyzed in these earlier studies, etc.) and how the current findings add to or are in contrast to those prior findings.

(4) The most interesting finding is buried on page 12 in the Discussion: "Uniquely, however, CD127 expression on CD4 T cells was significantly inversely associated with intact reservoir frequency." The authors should highlight this in the abstract, and title, and move this up in the Discussion. The paper describes a very high dimensional analysis and the key takeaways are not clear; the more the author can point the reader to the take-home points, the better their findings can have translatability to future follow-up mechanistic and/or validation studies.

(5) The authors should avoid overinterpretation of these results. For example in the Discussion on page 13 "The existence of two distinct clusters of PWH with different immune features and reservoir characteristics could have important implications for HIV cure strategies - these two groups may respond differently to a given approach, and cluster membership may need to be considered to optimize a given strategy." It is highly unlikely that future studies will be performing the breadth of parameters resulting here and then use these directly for optimizing therapy.

(6) There are only TWO limitations listed here: cross-sectional study design and the use of peripheral blood samples. (The subsequent paragraph notes an additional weakness which is misclassification of intact sequences by IPDA). This is a very limited discussion and highlights the need to more critically evaluate their study for potential weaknesses.

(7) A major clinical predictor of HIV reservoir size and decay is the timing of ART initiation. The authors should include these (as well as other clinical covariate data - see #12 below) in their analyses and/or describe as limitations of their study.

Reviewer #3 (Public Review):

Summary:

This valuable study by Semenova and colleagues describes a large cross-sectional cohort of 115 individuals on ART. Participants contributed a single blood sample which underwent IPDA, and 25-color flow with various markers (pre and post-stimulation). The authors then used clustering, decision tree analyses, and machine learning to look for correlations between these immunophenotypic markers and several measures of HIV reservoir volume. They identified two distinct clusters that can be somewhat differentiated based on total HIV DNA level, intact HIV DNA level, and multiple T cell cellular markers of activation and exhaustion.

The conclusions of the paper are supported by the data but the relationships between independent and dependent variables in the models are correlative with no mechanistic work to determine causality. It is unclear in most cases whether confounding variables could explain these correlations. If there is causality, then the data is not sufficient to infer directionality (ie does the immune environment impact the HIV reservoir or vice versa or both?). In addition, even with sophisticated and appropriate machine learning approaches, the models are not terribly predictive or highly correlated. For these reasons, the study is very much hypothesis-generating and will not impact cure strategies or HIV reservoir measurement strategies in the short term.

Strengths:

The study cohort is large and diverse in terms of key input variables such as age, gender, and duration of ART. Selection of immune assays is appropriate. The authors used a wide array of bioinformatic approaches to examine correlations in the data. The paper was generally well-written and appropriately referenced.

Weaknesses:

(1) The major limitation of this work is that it is highly exploratory and not hypothesis-driven. While some interesting correlations are identified, these are clearly hypothesis-generating based on the observational study design.

(2) The study's cross-sectional nature limits the ability to make mechanistic inferences about reservoir persistence. For instance, it would be very interesting to know whether the reservoir cluster is a feature of an individual throughout ART, or whether this outcome is dynamic over time.

(3) A fundamental issue is that I am concerned that binarizing the 3 reservoir metrics in a 50/50 fashion is for statistical convenience. First, by converting a continuous outcome into a simple binary outcome, the authors lose significant amounts of quantitative information. Second, the low and high reservoir outcomes are not actually demonstrated to be clinically meaningful: I presume that both contain many (?all) data points above levels where rebound would be expected soon after interruption of ART. Reservoir levels would also have no apparent outcome on the selection of cure approaches. Overall, dividing at the median seems biologically arbitrary to me.

(4) The two reservoir clusters are of potential interest as high total and intact with low % intact are discriminated somewhat by immune activation and exhaustion. This was the most interesting finding to me, but it is difficult to know whether this clustering is due to age, time on ART, other co-morbidity, ART adherence, or other possible unmeasured confounding variables.

(5) At the individual level, there is substantial overlap between clusters according to total, intact, and % intact between the clusters. Therefore, the claim in the discussion that these 2 cluster phenotypes may require different therapeutic approaches seems rather speculative. That said, the discussion is very thoughtful about how these 2 clusters may develop with consideration of the initial insult of untreated infection and / or differences in immune recovery.

(6) The authors state that the machine learning algorithms allow for reasonable prediction of reservoir volume. It is subjective, but to me, 70% accuracy is very low. This is not a disappointing finding per se. The authors did their best with the available data. It is informative that the machine learning algorithms cannot reliably discriminate reservoir volume despite substantial amounts of input data. This implies that either key explanatory variables were not included in the models (such as viral genotype, host immune phenotype, and comorbidities) or that the outcome for testing the models is not meaningful (which may be possible with an arbitrary 50/50 split in the data relative to median HIV DNA volumes: see above).

(7) The decision tree is innovative and a useful addition, but does not provide enough discriminatory information to imply causality, mechanism, or directionality in terms of whether the immune phenotype is impacting the reservoir or vice versa or both. Tree accuracy of 80% is marginal for a decision tool.

(8) Figure 2: this is not a weakness of the analysis but I have a question about interpretation. If total HIV DNA is more predictive of immune phenotype than intact HIV DNA, does this potentially implicate a prior high burden of viral replication (high viral load &/or more prolonged time off ART) rather than ongoing reservoir stimulation as a contributor to immune phenotype? A similar thought could be applied to the fact that clustering could only be detected when applied to total HIV DNA-associated features. Many investigators do not consider defective HIV DNA to be "part of the reservoir" so it is interesting to speculate why these defective viruses appear to have more correlation with immunophenotype than intact viruses.

(9) Overall, the authors need to do an even more careful job of emphasizing that these are all just correlations. For instance, HIV DNA cannot be proven to have a causal effect on the immunophenotype of the host with this study design. Similarly, immunophenotype may be affecting HIV DNA or the correlations between the two variables could be entirely due to a separate confounding variable.

(10) In general, in the intro, when the authors refer to the immune system, they do not consistently differentiate whether they are referring to the anti-HIV immune response, the reservoir itself, or both. More specifically, the sentence in the introduction listing various causes of immune activation should have citations. (To my knowledge, there is no study to date that definitively links proviral expression from reservoir cells in vivo to immune activation as it is next to impossible to remove the confounding possible imprint of previous HIV replication.) Similarly, it is worth mentioning that the depletion of intact proviruses is quite slow such that provial expression can only be stimulating the immune system at a low level. Similarly, the statement "Viral protein expression during therapy likely maintains antigen-specific cells of the adaptive immune system" seems hard to dissociate from the persistence of immune cells that were reactive to viremia.

(11) Given the many limitations of the study design and the inability of the models to discriminate reservoir volume and phenotype, the limitations section of the discussion seems rather brief.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

Semenova et al. have studied a large cross-sectional cohort of people living with HIV on suppressive ART, N=115, and performed high dimensional flow cytometry to then search for associations between immunological and clinical parameters and intact/total HIV DNA levels.

A number of interesting data science/ML approaches were explored on the data and the project seems a serious undertaking. However, like many other studies that have looked for these kinds of associations, there was not a very strong signal. Of course, the goal of unsupervised learning is to find new hypotheses that aren't obvious to human eyes, but I felt in that context, there were (1) results slightly oversold, (2) some questions about methodology in terms mostly of reservoir levels, and (3) results were not sufficiently translated back into meaning in terms of clinical outcomes.

We appreciate the reviewer’s perspective. In our revised version of the manuscript, we have attempted to address these concerns by more adequately explaining the limitations of the study and by more thoroughly discussing the context of the findings. We are not able to associate the findings with specific clinical outcomes for individual study participants but we speculate about the overall biological meaning of these associations across the cohort. We cannot disagree with the reviewer, but we find the associations statistically significant, potentially reflecting real biological associations, and forming the basis for future hypothesis testing research.

Strengths:

The study is evidently a large and impressive undertaking and combines many cutting-edge statistical techniques with a comprehensive experimental cohort of people living with HIV, notably inclusive of populations underrepresented in HIV science. A number of intriguing hypotheses are put forward that could be explored further. Sharing the data could create a useful repository for more specific analyses.

We thank the reviewer for this assessment.

Weaknesses:

Despite the detailed experiments and methods, there was not a very strong signal for the variable(s) predicting HIV reservoir size. The Spearman coefficients are ~0.3, (somewhat weak, and acknowledged as such) and predictive models reach 70-80% prediction levels, though sometimes categorical variables are challenging to interpret.

We agree with the reviewer that individual parameters are only weakly correlated with the HIV reservoir, likely reflecting the complex and multi-factorial nature of reservoir/immune cell interactions. Nevertheless, these associations are statistically significant and form the basis for functional testing in viral persistence.

There are some questions about methodology, as well as some conclusions that are not completely supported by results, or at minimum not sufficiently contextualized in terms of clinical significance. On associations: the false discovery rate correction was set at 5%, but data appear underdetermined with fewer observations than variables (144vars > 115ppts), and it isn't always clear if/when variables are related (e.g inverses of one another, for instance, %CD4 and %CD8).

When deriving a list of cell populations whose frequency would be correlated with the reservoir, we focused on well-defined cell types for which functional validation exists in the literature to consider them as distinct cell types. For many of the populations, gating based on combinations of multiple markers leads to recovery of very few cells, and so we excluded some potential combinations from the analysis. We are also making our raw data available for others to examine and find associations not considered by our manuscript.

The modeling of reservoir size was unusual, typically intact and defective HIV DNA are analyzed on a log10 scale (both for decays and predicting rebound). Also, sometimes in this analysis levels are normalized (presumably to max/min?, e.g. S5), and given the large within-host variation of level we see in other works, it is not trivial to predict any downstream impact of normalization across population vs within-person.

We have repeated the analysis using log10 transformed data and the new figures are shown in Figure 1 and S2-S5.

Also, the qualitative characterization of low/high reservoir is not standard and naturally will split by early/later ART if done as above/below median. Given the continuous nature of these data, it seems throughout that predicting above/below median is a little hard to translate into clinical meaning.

Our ML models included time before ART as a variable in the analysis, and this was not found to be a significant driver of the reservoir size associations, except for the percentage of intact proviruses (see Figure 2C). Furthermore, we analyzed whether any of the reservoir correlated immune variables were associated with time on ART and found that, although some immune variables are associated with time on therapy, this was not the case for most of them (Table S4). We agree that it is challenging to translate above or below median into clinical meaning for this cohort, but we emphasize that this study is primarily a hypothesis generating approach requiring additional validation for the associations observed. We attempted to predict reservoir size as a continuous variable using the data and this approach was not successful (Figure S13). We believe that a significantly larger cohort will likely be required to generate a ML model that can accurately predict the reservoir as a continuous variable. We have added additional discussion of this to the manuscript.

Lastly, the work is comprehensive and appears solid, but the code was not shared to see how calculations were performed.

We now provide a link to the code used to perform the analyses in the manuscript, https://github.com/lesiasemenova/ML_HIV_reservoir.

Reviewer #2 (Public Review):

Summary:

Semenova et. al., performed a cross-sectional analysis of host immunophenotypes (using flow cytometry) and the peripheral CD4+ T cell HIV reservoir size (using the Intact Proviral DNA Assay, IPDA) from 115 people with HIV (PWH) on ART. The study mostly highlights the machine learning methods applied to these host and viral reservoir datasets but fails to interpret these complex analyses into (clinically, biologically) interpretable findings. For these reasons, the direct translational take-home message from this work is lost amidst a large list of findings (shown as clusters of associated markers) and sentences such as "this study highlights the utility of machine learning approaches to identify otherwise imperceptible global patterns" - lead to overinterpretation of their data.

We have addressed the reviewer’s concern by modifications to the manuscript that enhance the interpretation of the findings in a clinical and biological context.

Strengths:

Measurement of host immunophenotyping measures (multiparameter flow cytometry) and peripheral HIV reservoir size (IPDA) from 115 PWH on ART.

Major Weaknesses:

(1) Overall, there is little to no interpretability of their machine learning analyses; findings appear as a "laundry list" of parameters with no interpretation of the estimated effect size and directionality of the observed associations. For example, Figure 2 might actually give an interpretation of each X increase in immunophenotyping parameter, we saw a Y increase/decrease in HIV reservoir measure.

We have added additional text to the manuscript in which we attempt to provide more immunological and clinical interpretation of the associations. We also have emphasized that these associations are still speculative and will require additional validation. Nevertheless, our data should provide a rich source of new hypotheses regarding immune system/reservoir interaction that could be tested in future work.

(2) The correlations all appear to be relatively weak, with most Spearman R in the 0.30 range or so.

We agree with the review that the associations are mostly weak, consistent with previous studies in this area. This likely is an inherent feature of the underlying biology – the reservoir is likely associated with the immune system in complex ways and involves stochastic processes that will limit the predictability of reservoir size using any single immune parameter. We have added additional text to the manuscript to make this point clearer.

(3) The Discussion needs further work to help guide the reader. The sentence: "The correlative results from this present study corroborate many of these studies, and provide additional insights" is broad. The authors should spend some time here to clearly describe the prior literature (e.g., describe the strength and direction of the association observed in prior work linking PD-1 and HIV reservoir size, as well as specify which type of HIV reservoir measures were analyzed in these earlier studies, etc.) and how the current findings add to or are in contrast to those prior findings.

We have added additional text to the manuscript to help guide the readers through the possible biological significance of the findings and the context with respect to prior literature.

(4) The most interesting finding is buried on page 12 in the Discussion: "Uniquely, however, CD127 expression on CD4 T cells was significantly inversely associated with intact reservoir frequency." The authors should highlight this in the abstract, and title, and move this up in the Discussion. The paper describes a very high dimensional analysis and the key takeaways are not clear; the more the author can point the reader to the take-home points, the better their findings can have translatability to future follow-up mechanistic and/or validation studies.

We appreciate the reviewer’s comment. We have increased the emphasis on this finding in the revised version of the manuscript.

(5) The authors should avoid overinterpretation of these results. For example in the Discussion on page 13 "The existence of two distinct clusters of PWH with different immune features and reservoir characteristics could have important implications for HIV cure strategies - these two groups may respond differently to a given approach, and cluster membership may need to be considered to optimize a given strategy." It is highly unlikely that future studies will be performing the breadth of parameters resulting here and then use these directly for optimizing therapy.

Our analyses indicate that membership of study participants in cluster1 or cluster 2 can be fairly accurately determined by a small number of individual parameters (KLRG1 etc, Figure 4F), and measuring the cells of PWH with the degree of breadth used in this paper would not be necessary to classify PWH into these clusters. As such, we feel that it is not unrealistic to speculate that this finding could turn out to be clinically useful, if it becomes clear that the clusters are biologically meaningful.

(6) There are only TWO limitations listed here: cross-sectional study design and the use of peripheral blood samples. (The subsequent paragraph notes an additional weakness which is misclassification of intact sequences by IPDA). This is a very limited discussion and highlights the need to more critically evaluate their study for potential weaknesses.

We have expanded on the list of limitations discussed in the manuscript. In particular, we now address the size of the cohort, the composition with respect to different genders and demographics, lack of information for the timing of ART and the lack of information regarding intracellular transcriptional pathways.

(7) A major clinical predictor of HIV reservoir size and decay is the timing of ART initiation. The authors should include these (as well as other clinical covariate data - see #12 below) in their analyses and/or describe as limitations of their study.

All of the participants that make up our cohort were treated during chronic infection, and the precise timing of ART initiation is unclear in most of these cases. We have added additional information to explain this in the manuscript and include this in the list of limitations.

Reviewer #3 (Public Review):

Summary:

This valuable study by Semenova and colleagues describes a large cross-sectional cohort of 115 individuals on ART. Participants contributed a single blood sample which underwent IPDA, and 25-color flow with various markers (pre and post-stimulation). The authors then used clustering, decision tree analyses, and machine learning to look for correlations between these immunophenotypic markers and several measures of HIV reservoir volume. They identified two distinct clusters that can be somewhat differentiated based on total HIV DNA level, intact HIV DNA level, and multiple T cell cellular markers of activation and exhaustion.

The conclusions of the paper are supported by the data but the relationships between independent and dependent variables in the models are correlative with no mechanistic work to determine causality. It is unclear in most cases whether confounding variables could explain these correlations. If there is causality, then the data is not sufficient to infer directionality (ie does the immune environment impact the HIV reservoir or vice versa or both?). In addition, even with sophisticated and appropriate machine learning approaches, the models are not terribly predictive or highly correlated. For these reasons, the study is very much hypothesis-generating and will not impact cure strategies or HIV reservoir measurement strategies in the short term.

We appreciate the reviewer’s comments regarding the value of our study. We fully acknowledge that the causal nature and directionality of these associations are not yet clear and agree that the study is primarily hypothesis generating in nature. Nevertheless, we feel that the hypotheses generated will be valuable to the field. We have added additional text to the manuscript to emphasize the hypothesis generating nature of this paper.

Strengths:

The study cohort is large and diverse in terms of key input variables such as age, gender, and duration of ART. Selection of immune assays is appropriate. The authors used a wide array of bioinformatic approaches to examine correlations in the data. The paper was generally well-written and appropriately referenced.

Weaknesses:

(1) The major limitation of this work is that it is highly exploratory and not hypothesis-driven. While some interesting correlations are identified, these are clearly hypothesis-generating based on the observational study design.

We agree that the major goal of this study was hypothesis generating and that our work is exploratory in nature. Performing experiments with mechanism testing goals in human participants with HIV is challenging. Additionally, before such mechanistic studies can be undertaken, one must have hypotheses to test. As such we feel our study will be useful for the field in helping to identify hypotheses that could potentially be tested.

(2) The study's cross-sectional nature limits the ability to make mechanistic inferences about reservoir persistence. For instance, it would be very interesting to know whether the reservoir cluster is a feature of an individual throughout ART, or whether this outcome is dynamic over time.

We agree with the reviewer’s comment. Longitudinal studies are challenging to carry out with a study cohort of this size, and addressing questions such as the one raised by the reviewer would be of great interest. We believe our study nevertheless has value in identifying hypotheses that could be tested in a longitudinal study.

(3) A fundamental issue is that I am concerned that binarizing the 3 reservoir metrics in a 50/50 fashion is for statistical convenience. First, by converting a continuous outcome into a simple binary outcome, the authors lose significant amounts of quantitative information. Second, the low and high reservoir outcomes are not actually demonstrated to be clinically meaningful: I presume that both contain many (?all) data points above levels where rebound would be expected soon after interruption of ART. Reservoir levels would also have no apparent outcome on the selection of cure approaches. Overall, dividing at the median seems biologically arbitrary to me.

The reviewer raises a valid point that the clinical significance of above or below median reservoir metrics is unclear, and that the size of the reservoir has potentially little relation to rebound and cure approaches. In the manuscript, we attempted to generate models that can predict reservoir size as a continuous variable in Figure S13 and find that this approach performs poorly, while a binarized approach was more successful. As such we have included both approaches in the manuscript. It is possible that future studies with larger sample sizes and more detailed measurements will perform better for continuous variable prediction. While this is a fairly large study (n=115) by the standards of HIV reservoir analyses, it is a small study by the standards of the machine learning field, and accurate predictive ML models for reservoir size as a continuous variable will likely require a much larger set of samples/participants. Nevertheless, we feel our work has value as a template for ML approaches that may be informative for understanding HIV/immune interactions and generates novel hypotheses that could be validated by subsequent studies.

(4) The two reservoir clusters are of potential interest as high total and intact with low % intact are discriminated somewhat by immune activation and exhaustion. This was the most interesting finding to me, but it is difficult to know whether this clustering is due to age, time on ART, other co-morbidity, ART adherence, or other possible unmeasured confounding variables.

We agree that this finding is one of the more interesting outcomes of the study. We examined a number of these variables for association with cluster membership, and these data are reported in Figure S8A-D. Age, years of ART and CD4 Nadir were all clearly different between the clusters. The striking feature of this clustering, however, is the clear separation between the two groups of participants, as opposed to a continuous gradient of phenotypes. This could reflect a bifurcation of outcomes for people with HIV, dynamic changes in the reservoir immune interactions over time, or different levels of untreated infection. It is certainly possible that some other unmeasured confounding variables contribute to this outcome and we have attempted to make this limitation clearer.

(5) At the individual level, there is substantial overlap between clusters according to total, intact, and % intact between the clusters. Therefore, the claim in the discussion that these 2 cluster phenotypes may require different therapeutic approaches seems rather speculative. That said, the discussion is very thoughtful about how these 2 clusters may develop with consideration of the initial insult of untreated infection and / or differences in immune recovery.

We agree with the reviewer that this claim is speculative, and we have attempted to moderate the language of the text in the revised version.

(6) The authors state that the machine learning algorithms allow for reasonable prediction of reservoir volume. It is subjective, but to me, 70% accuracy is very low. This is not a disappointing finding per se. The authors did their best with the available data. It is informative that the machine learning algorithms cannot reliably discriminate reservoir volume despite substantial amounts of input data. This implies that either key explanatory variables were not included in the models (such as viral genotype, host immune phenotype, and comorbidities) or that the outcome for testing the models is not meaningful (which may be possible with an arbitrary 50/50 split in the data relative to median HIV DNA volumes: see above).

We acknowledge that the predictive power of the models generated from these data is modest and we have clarified this point in the revised manuscript. As the reviewer indicates, this may result from the influence of unmeasured variables and possible stochastic processes. The data may thus demonstrate a limit to the predictability of reservoir size which may be inherent to the underlying biology. As we mention above, this study size (n-115) is fairly small for the application of ML methods, and an increased sample size will likely improve the accuracy of the models. At this stage, the models we describe are not yet useful as predictive clinical tools, but are still nonetheless useful as tools to describe the structure of the data and identify reservoir associated immune cell types.

(7) The decision tree is innovative and a useful addition, but does not provide enough discriminatory information to imply causality, mechanism, or directionality in terms of whether the immune phenotype is impacting the reservoir or vice versa or both. Tree accuracy of 80% is marginal for a decision tool.

The reviewer is correct about these points. In the revised manuscript, we have attempted to make it clear that we are not yet advocating using this approach as a decision tool, but simply a way to visualize the data and understand the structure of the dataset. As we discuss above, the models will likely need to be trained on a larger dataset and achieve higher accuracy before use as a decision tool.

(8) Figure 2: this is not a weakness of the analysis but I have a question about interpretation. If total HIV DNA is more predictive of immune phenotype than intact HIV DNA, does this potentially implicate a prior high burden of viral replication (high viral load &/or more prolonged time off ART) rather than ongoing reservoir stimulation as a contributor to immune phenotype? A similar thought could be applied to the fact that clustering could only be detected when applied to total HIV DNA-associated features. Many investigators do not consider defective HIV DNA to be "part of the reservoir" so it is interesting to speculate why these defective viruses appear to have more correlation with immunophenotype than intact viruses.

We agree with the reviewer that this observation could reflect prior viral burden and we have added additional text to make this clearer. Even so, we cannot rule out a model in which defective viral DNA is engaged in ongoing stimulation of the immune system during ART, leading to the stronger association between total DNA and the immune cell phenotypes. We hypothesize that the defective proviruses could potentially be triggering innate immune pattern recognition receptors via viral RNA or DNA, and a higher burden of the total reservoir leads to a stronger apparent association with the immune phenotype. We have included text in the discussion about this hypothesis.

(9) Overall, the authors need to do an even more careful job of emphasizing that these are all just correlations. For instance, HIV DNA cannot be proven to have a causal effect on the immunophenotype of the host with this study design. Similarly, immunophenotype may be affecting HIV DNA or the correlations between the two variables could be entirely due to a separate confounding variable

We have revised the text of the manuscript to emphasize this point, and we acknowledge that any causal relationships are, at this point, simply speculation.

(10) In general, in the intro, when the authors refer to the immune system, they do not consistently differentiate whether they are referring to the anti-HIV immune response, the reservoir itself, or both. More specifically, the sentence in the introduction listing various causes of immune activation should have citations. (To my knowledge, there is no study to date that definitively links proviral expression from reservoir cells in vivo to immune activation as it is next to impossible to remove the confounding possible imprint of previous HIV replication.) Similarly, it is worth mentioning that the depletion of intact proviruses is quite slow such that provial expression can only be stimulating the immune system at a low level. Similarly, the statement "Viral protein expression during therapy likely maintains antigen-specific cells of the adaptive immune system" seems hard to dissociate from the persistence of immune cells that were reactive to viremia.

We updated the text of the manuscript to address these points and have added additional citations as per the reviewer’s suggestion.

(11) Given the many limitations of the study design and the inability of the models to discriminate reservoir volume and phenotype, the limitations section of the discussion seems rather brief.

We have now expanded the limitations section of the discussion and added additional considerations. We now include a discussion of the study cohort size, composition and the detail provided by the assays.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

A few specific comments:

"This pattern is likely indicative of a more profound association of total HIV DNA with host immunophenotype relative to intact HIV DNA."

Most studies I have seen (e.g. single cell from Lictherfeld/Yu group) show intact proviruses are generally more activated/detectable/susceptible to immune selection, so I have a hard time thinking defective proviruses are actually more affected by immunotype.

We hypothesize that this association is actually occurring in the opposite direction – that the defective provirus are having a greater impact on the immune phenotype, due to their greater number and potential ability to engage innate or adaptive immune receptors. We have clarified this point in the manuscript

"The existence of two distinct clusters of PWH with different immune features and reservoir characteristics could have important implications for HIV cure strategies - these two groups may respond differently to a given approach, and cluster membership may need to be considered to optimize a given strategy."

I find this a bit of a reach, given that the definition of 2 categories depended on the total size.

We have modified the language of this section to reduce the level of speculation.

"This study is cross-sectional in nature and is primarily observational, so caution should be used interpreting findings associated with time on therapy".

I found this an interesting statement because ultimately time on ART shows up throughout the analysis as a significant predictor, do you mean something about how time on ART could indicate other confounding variables like ART regimen or something?

We have rephrased this comment to avoid confusion. We were simply trying to make the point that we should avoid speculating about longitudinal dynamics from cross sectional data.

"As expected, the plots showed no significant correlation for intact HIV DNA versus years of ART (Figure 1B), while total reservoir size was positively correlated with the time of ART (Figure 1A, Spearman r = 0.31)."

Is this expected? Studies with longitudinal data almost uniformly show intact decay, at least for the first 10 or so years of ART, and defective/total stability (or slight decay). Also probably "time on ART" to not confuse with the duration of infection before ART.

We have updated the language of this section to address this comment. We have avoided comparing our data with respect to time on ART to longitudinal studies for reasons given above.

On dimensionality reduction, as this PaCMAP seems a relatively new technique (vs tSNE and UMAP which are more standard, but absolutely have their weaknesses), it does seem important to contextualize. I think it would still be useful to show PCA and asses the % variance of each additional dimension to assess the effective dimensionality, it would be helpful to show a plot of % variance by # components to see if there is a cutoff somewhere, and if PaCMAP is really picking this up to determine the 2 dimensions/2 clusters is ideal. Figure 4B ultimately shows a lot of low/high across those clusters, and since low/high is defined categorically it's hard to know which of those dots are very close to the other categories.

We have added this analysis to the manuscript – found in Figure S9. The PCA plot indicates that members of the two clusters also separate on PCA although this separation is not as clear as for the PaCMAP plot.

Minor comments on writing etc:

Intro

-Needs some references on immune activation sequelae paragraph.

We have added some additional references to this section.

-"promote the entry of recently infected cells into the reservoir" -- that is only one possible mechanistic explanation, it's not unreasonable but it seems important to keep options open until we have more precise data that can illuminate the mechanism of the overabundance.

We have modified the text to discuss additional hypotheses.

-You might also reference Pankau et al Ppath for viral seeding near the time of ART.

We have added this reference.

-"Viral protein expression during therapy likely maintains antigen-specific cells of the adaptive immune system" - this was unclear to me, do you mean HIV-specific cells that act against HIV during ART? I think most studies show immunity against HIV (CD8 and CD4) wanes over time during ART.

The Goonetilleke lab has recently generated data indicating that antiviral T cell responses are remarkably stable over time on ART, but we agree with the reviewer that the idea that ongoing antigen expression in the reservoir maintains these cells is speculative. We have modified the text to make this point clearer.

-Overall I think the introduction lacked a little bit of definitional precision: i.e. is the reservoir intact vs replication competent vs all HIV DNA and whether we are talking about PWH on long-term ART and how long we should be imagining? The first years of ART are certainly different than later, in terms of dynamics. The ultimate implications are likely specific for some of these categorizations.

-"persistent sequelae of the massive disruptions to T cell homeostasis and lymphoid structures that occur during untreated HIV infection" needs a lot more context/referencing. For instance, Peter Hunt showed a decrease in activation after ART a long time ago.

-Heather Best et al show T cell clonality stays perturbed after ART.

We have updated the text of the introduction and added references to address the reviewer’s comments.

Results

-It would be important to mention the race of participants and any information about expected clades of acquired viruses, this gets mentioned eventually with reference to the Table but the breakdown would be helpful right away.

We have added this information to the results section.

-"performed Spearman correlations", may be calculated or tested?

We have corrected the language for this sentence.

Comments on figures:

-Figure 1 data on linear scale (re discussion above) -- hard to even tell if there is a decay (to match with all we know from various long-term ART studies).

-Figure 4 data is shown on ln (log_e) scale, which is hard to interpret for most people.

-Figures 4 C,D, and E should have box plots to visually assess the significance.

-Figure 4B legend says purple/pink but I think the colors are different in the plot, could be about transparency

-Figure 5 it is now not clear if log_e(?).

-Figure 6 "HIV reservoir characteristics" might be better to make this more explicit. Do you mean for instance in the 6B title Total HIV DNA per million CD4+ T cells I think?

We have made these modifications.

Reviewer #2 (Recommendations For The Authors):

Minor Weaknesses:

(1) The Introduction is too long and much of the text is not directly related to the study's research question and design.

We have streamlined the introduction in the revised manuscript.

(2) While no differences were seen by age or race, according to the authors, this is unlikely to be useful since the numbers are so small in some of these subcategories. Results from sensitivity analyses (e.g., excluding these individuals) may be more informative/useful.

We agree that the lower numbers of participants for some subgroupings makes it challenging to know for sure if there are any differences based on these variables. Have added text to clarify this. We have added age, race and gender to the LOCO analysis and to the variable inflation importance analysis (Table S5).

(3) For Figure 4, based on what was described in the Results section of the manuscript, the authors should clarify that the figures show results for TOTAL HIV DNA only (not intact DNA): "Dimension reduction machine learning approaches identified two robust clusters of PWH when using total HIV DNA reservoir-associated immune cell frequencies (Figure 4A), but not for intact or percentage intact HIV DNA (Figure 4B and 4C)".

We have added this information.

(4) The statement on page 5, first paragraph, "Interestingly, when we examined a plot of percent intact proviruses versus time on therapy (Figure 1C), we observed a biphasic decay pattern," is not new (Peluso JCI Insight 2020, Gandhi JID 2023, McMyn JCI 2023). Prior studies have clearly demonstrated this biphasic pattern and should be cited here, and the sentence should be reworded with something like "consistent with prior work", etc.

We have added citations to these studies and rephrased this comment.

(5) The Cohort and sample collection sections are somewhat thin. Further details on the cohort details should include at the very minimum some description of the timing of ART initiation (is this mostly a chronic-treated cohort?) and important covariate data such as nadir CD4+ T cell count, pre-ART viral load, duration of ART suppression, etc.

The cohort was treated during chronic infection, and we have clarified this in the manuscript. Information regarding CD4 nadir and years on ART are included in Table 1. Unfortunately, pre-ART viral load was not available for most members of this cohort, so we did not use it for analyses. The partial pre-ART viral load data is included with the dataset we are making publicly available.

Reviewer #3 (Recommendations For The Authors):

Minor points:

(1) What is meant by CD4 nadir? Is this during primary infection or the time before ART initiation?

We have clarified this description in the manuscript. This term refers to the lowest CD4 count recorded during untreated infection.

(2) The authors claim that determinants of reservoir size are starting to emerge but other than the timing of ART, I am not sure what studies they are referring to.

We have updated the language of this section. We intended to refer to studies looking at correlates of reservoir size, and feel that this is a more appropriate term that ‘determinants’

(3) The discussion does not tie in the model-generated hypotheses with the known mechanisms that sustain the reservoir: clonal proliferation balanced by death and subset differentiation. It would be interesting to tie in the proposed reservoir clusters with these known mechanisms.

We have added additional text to the manuscript to address these mechanisms.

(4) Figure 1: Total should be listed as total HIV DNA.

We have updated this in the manuscript.

(5) Figure 1C: Worth mentioning the paper by Reeves et al which raises the possibility that the flattening of intact HIV DNA at 9 years may be spurious due to small levels of misclassification of defective as intact.

We have added this reference.

(6) "Total reservoir frequency" should be "total HIV DNA concentration"

We respectfully feel that “frequency” is a more accurate term than “concentration”, since we are expressing the reservoir as a fraction of the CD4 T cells, while “concentration” suggests a denominator of volume.

(7) Figure S2-5: label y-axis total HIV DNA.

We have updated this figure.

Version 2

Abstract

Abstract

Understanding the interplay between the HIV reservoir and the host immune system may yield insights into HIV persistence during antiretroviral therapy (ART) and inform strategies for a cure. Here, we applied machine learning approaches to cross-sectional high-parameter HIV reservoir and immunology data in order to characterize host-reservoir associations and generate new hypotheses about HIV reservoir biology. High-dimensional immunophenotyping, quantification of HIV-specific T cell responses, and measurement of genetically intact and total HIV proviral DNA frequencies were performed on peripheral blood samples from 115 people with HIV (PWH) on long-term ART. Analysis demonstrated that both intact and total proviral DNA frequencies were positively correlated with T cell activation and exhaustion. Years of ART and select bifunctional HIV-specific CD4 T cell responses were negatively correlated with the percentage of intact proviruses. A Leave-One-Covariate-Out (LOCO) inference approach identified specific HIV reservoir and clinical-demographic parameters, such as age and biological sex, that were particularly important in predicting immunophenotypes. Overall, immune parameters were more strongly associated with total HIV proviral frequencies than intact proviral frequencies. Uniquely, however, expression of the IL-7 receptor alpha chain (CD127) on CD4 T cells was more strongly correlated with the intact reservoir. Unsupervised dimension reduction analysis identified two main clusters of PWH with distinct immune and reservoir characteristics. Using reservoir correlates identified in these initial analyses, decision tree methods were employed to visualize relationships among multiple immune and clinical-demographic parameters and the HIV reservoir. Finally, using random splits of our data as training-test sets, machine learning algorithms predicted with approximately 70% accuracy whether a given participant had qualitatively high or low levels of total or intact HIV DNA. The techniques described here may be useful for assessing global patterns within the increasingly high-dimensional data used in HIV reservoir and other studies of complex biology.

Introduction

The advent of ART has significantly decreased morbidity and mortality associated with HIV infection. However, a long-lived proviral reservoir precludes a cure of infection (1–4). While a number of cell types have been proposed as contributing to the HIV reservoir, the largest and most well-characterized reservoir is within CD4 T cells. Molecular analyses of residual HIV proviruses in CD4 T cells during ART have shown that the reservoir is comprised of a heterogeneous mixture of full-length replication-competent proviruses and a larger set of defective proviruses containing internal deletions or APOBEC-mediated hypermutation(s) (5–7). Infected cells can undergo clonal expansion and contraction over time in response to homeostatic cues or antigen-driven activation (8–10). In addition to the presence of a latent HIV reservoir, PWH on therapy exhibit a persistent level of T cell immune activation and other altered phenotypes compared to uninfected individuals, involving cells of both the innate and adaptive immune systems (11). Specifically, HIV infection is associated with increased levels of inflammatory biomarkers and increased immune cell activation (12–15). PWH also exhibit a depleted naïve T cell compartment (16) and increased expression of activation/exhaustion and senescence markers such as CD38, PD-1 and KLRG1(17–21). The molecular mechanisms underlying persistent immune activation and dysfunction for PWH on ART are unclear, but it is possible that ongoing viral gene expression and reverse transcription generates ligands for pattern recognition receptors in the innate immune system (22). However, another important potential cause of ongoing immune activation during ART is persistent sequelae of the massive disruptions to T cell homeostasis, intestinal permeability and lymphoid structures that occur during untreated HIV infection (23–25).

It is possible that the HIV reservoir and the host immune system interact in multiple complex ways during therapy. The dynamics of how the HIV reservoir affects the immune system and vice versa are important to characterize and may lead to novel ideas to promote reservoir depletion or to inhibit immune dysfunction that may be related to ongoing HIV expression in persistently infected cells. A number of previous studies have examined the relationship between reservoir size and the immune system (21,26–44). Consistent major findings from these studies include enrichment of the HIV reservoir in CD4 T cells expressing activation and exhaustion markers, positive correlation of the reservoir with central memory T cells (Tcm) and transitional memory (Ttm) T cell subsets, and inverse correlation of the HIV reservoir with the CD4 nadir and the CD4/CD8 T cell ratio. Recent studies indicate enrichment of the intact HIV reservoir in cells expressing VLA-4, but no correlation with classic T cell activation markers (26, 36). One important limitation of most of these studies is that they were not able to distinguish intact proviruses from defective proviruses, making the relationship between the immune system and the intact reservoir unclear. Furthermore, most of these studies examined a limited number of surface markers and may have missed associations that would be revealed by a higher-resolution examination of immunophenotypes. Additional assessment of immune correlates of the intact HIV reservoir in additional cohorts is thus needed. Finally, there is a recent, growing appreciation for the profound influence of ART initiation timing and length of treatment on the size and composition of the HIV reservoir that must be accounted for in assessing persistent immune activation in PWH on therapy (11, 45).

In this study, we use a mixture of traditional statistical approaches and novel data science methods to assess the relationship between the immune system and the HIV reservoir across a cross-sectional study of 115 PWH. We identify numerous significant correlations between immune cell populations and the size of the total HIV reservoir, as well as smaller number of associations with the intact reservoir, including CD127 expression in CD4 T cells. This study highlights the potential of machine learning approaches to visualize global patterns in high-parameter studies of the HIV reservoir. Additionally, these approaches corroborate recent findings in the HIV persistence field regarding preferential intact proviral decay and immune dynamics and highlight the complex immunologic signatures of HIV latency.

Results

Cohort description, reservoir quantification, and immunophenotyping

This cross-sectional study evaluated 115 participants with HIV infection for at least one year and who had been receiving ART for at least 0.9 years (median 9 years, IQR 5.2-16.6) (Table 1). The study cohort was 77 percent male, 60 percent black and median participant age was 45 years.

The Intact Proviral DNA assay (IPDA) was performed on isolated total CD4 T cells (46). The IPDA estimates of the frequency of total proviruses, intact proviruses, and the percent intact of total proviruses for each participant. Immunophenotyping was performed using 25-color spectral flow cytometry on peripheral blood mononuclear cells (PBMCs) from each study participant. A representative flow gating is shown in Figure S1. Briefly, cells within the live lymphocytes gate were first defined as T cells (CD3+/CD56-), NK cells (CD3-/CD56+), or NKT cells (CD3+/CD56+). T cells were then further subdivided as CD4+ or CD8+ T cells, and then as naïve T cells (Tn, CD45RA+/CCR7+), central memory T cells (Tcm, CD45RA-/CCR7+), effector memory cells (Tem, CD45RA-/CCR7-) and terminally differentiated effector cells (Teff, CD45RA+/CCR7-). Expression of various surface markers within each of these subsets was then examined (CD38, HLA-DR, PD-1, KLRG1, CD127, CD27, NKG2A). In addition to measuring the baseline abundance of surface proteins, we also stimulated PBMCs from each participant with or without a pool of HIV-derived peptides and measured intracellular cytokines TNFα, IL-2, and IFNγ as well as the surface expression of the degranulation marker CD107a. HIV-specific T cell responses were defined as the difference in frequency of cytokine/CD107a+ cells between HIV stimulated and nonstimuated cells. Frequencies of CD4 or CD8 T cells that were singleor double-positive for TNFα, IL-2, IFNγ, or CD107a were used for analysis.

These analyses resulted in the determination of a total of 144 parameters for the cohort of n = 115 PWH on ART. The 144 parameters include 133 immunophenotypic cell frequencies, 3 HIV reservoir parameters (frequency of intact HIV DNA, frequency of total HIV DNA, and percentage intact of total HIV DNA), and 8 clinical-demographic parameters. Clinical-demographic variables (Table 1) included age, biological sex, years of ART, estimated years of HIV infection prior to ART (here, we use categorical variables: years before ART=NA, years before ART< 1, years before ART≥ 1), CD4 nadir, most recent CD4 T cell count, and race (Caucasian, African-American, other). No significant differences in the reservoir based on biological sex or race were observed (Table S1). The median and interquartile range (IQR) for all immune cells population variables are presented in Table S2.

Correlation of total, intact, and percent intact HIV DNA with immunophenotypes

For the initial analysis, Spearman correlation of the clinical-demographic and immunophenotypic data with HIV reservoir metrics was assessed. 69 host variables correlated (unadjusted p < 0.05) with one or more characteristics of the HIV reservoir (Table S3). Following Benjamini-Hochberg p-value adjustment for multiple comparisons using a false discovery rate correction at 5%, 31 variables were found to be correlated with total HIV DNA, 3 with intact HIV DNA, and 4 with the percentage of intact HIV DNA (Table 2). Of note, the correlations observed were generally weak to moderate in magnitude (the mean of the absolute values of Spearman correlation coefficients was 0.31 for total HIV DNA, 0.33 for intact HIV DNA, and 0.35 for percent intact HIV DNA).

With regard to clinical-demographic variables, CD4 nadir prior to ART initiation was negatively correlated with total HIV DNA (Spearman r = −0.32), whereas age (r = 0.31) and years of ART (r = 0.31) were positively correlated with total HIV DNA (Table 2). Years of ART (r = −0.45) was negatively correlated with the percentage of intact proviruses (Table 2), consistent with the hypothesis that intact proviruses are progressively eliminated from the reservoir over time. We then examined the association between the reservoir and time on ART further (Figure 1). Correlation plots of total reservoir frequency, intact reservoir frequency, and percentage intact HIV DNA versus years of ART were visualized (Figure 1A-C). The plots showed no significant correlation for intact HIV DNA frequency versus years of ART (Figure 1B), while total HIV DNA reservoir frequency was positively correlated with the length of ART treatment. (Figure 1A, Spearman r = 0.31). When we examined a plot of percent intact proviruses versus time on therapy (Figure 1C), we observed a biphasic decay pattern, consistent with previous reports (47–49). Indeed, when we fitted piece-wise linear functions with allowances for up to 2 breaks, the best-fit model had decay slope changes at 1.6 and 9 years of ART. The R2 score of a piece-wise linear model with two breaks was 44.18%, while the score for a linear model without any breaks was 16.3%. In PWH within approximately 0-6 years of ART initiation, a significant proportion of the participants exhibited high fractions of intact proviruses (50-90%). After approximately 6 years of ART, however, the majority of participants had reservoirs for which intact proviruses represented a minor fraction of the overall reservoir (0-30%). Of note, several immune features also demonstrated statistically significant correlations with years of ART, including CD8 T cell expression of CD38 (negative correlation), CD8 Tcm frequency (positive correlation), and CD107a+IL-2+IFNγ-TNFα-CD8 T cells (positive correlation)(Figures S2, S3, S4, Table S4).

Duration of treatment and the HIV reservoir.

Scatterplots for years of ART vs total HIV reservoir frequency (A), intact reservoir frequency (B) and percent intact (C) are shown. Each dot represents an individual study participant. Correlation coefficients and corresponding p-values are shown for each plot. Participants that have missing values of years of ART were not included in the plot. For percent intact, piece-wise linear function with two breaks is fitted. For total HIV reservoir frequency a linear function is fitted.

Examination of immunophenotypic parameter correlations with HIV DNA metrics demonstrated several notable correlations (Table 2, Figures S2, S3, S4). As previously reported, CD4 T cell frequencies were negatively correlated with total HIV DNA frequencies (r = −0.36), whereas CD8 T cells were pos-itively correlated (r = 0.41). This relationship held true for intact HIV DNA frequencies but not for the percentage of intact HIV DNA (Table 2). Frequencies of CD4 and CD8 T naïve cells, as well as their expression of CD38, were significantly negatively correlated with total HIV DNA frequency, consistent with the hypothesis that larger reservoirs are associated with depletion of the naive T cell subsets and increased immune activation. By contrast, CD4 and CD8 Tcm and Tem cells were weakly positively correlated with total HIV DNA, possibly due to increased differentiation of the naive compartment into these memory subsets in people with larger reservoirs. Markers of T cell activation and exhaustion were weakly but significantly positively correlated with total HIV DNA, including NKG2A+ CD4 T, PD-1+ CD4 T, HLA-DR+ CD4 T, PD-1+ CD8 T, and PD-1+/CCR7+ CD8 T frequencies, among others (Table 2, Figures S2, S3, S4). For both intact and total HIV DNA, an inverse correlation with the CD4/CD8 T cell ratio was observed (Figure S5). Interestingly, for the percentage of intact HIV DNA, negative correlations were observed with the frequencies of 2 subsets of bifunctional HIV peptide-stimulated CD4 T cells (IL-2+TNFα+ and IFNγ+IL-2+), as well as with years on ART. Notably, the magnitude of correlation with immune parameters was typically stronger for total HIV DNA compared to intact HIV DNA (Table 2). This pattern may reflect a larger influence of total HIV DNA on immunophenotype of the host, due to the overall higher abundance of total HIV DNA relative to the intact reservoir. A notable exception to this pattern was CD127-CD4 T cell frequencies, which were significantly positively correlated with intact but not total HIV DNA.

Overall these findings support the notion that larger HIV reservoirs are associated with increased levels of immune activation and depletion of the naive T cell compartment. Furthermore, the data indicate that there may be some immune features, such as CD127 expression on CD4 T cells that are uniquely associated with the intact reservoir rather than the total reservoir.

Leave-One-Covariate-Out (LOCO) inference identifies specific HIV reservoir and clinical-demographic parameters important for the prediction of immunophenotypes

Given the evidence that reservoir and immune recovery dynamics likely occur in concert, we next used a variable importance approach - Leave-One-Covariate-Out (LOCO) inference analysis, to account for potential confounding variables and more carefully assess the relative importance of a given variable to an immunophenotype. The LOCO inference approach is described in Figure 2A. First, a least-squares regression linear model that predicts the dependent variable is fitted on a set of independent variables, and an R2 value is generated (Figure 2A Step 1). Next, one independent variable is excluded from the model, the model is refitted, and an adjusted R2 value is determined (Figure 2A Step 2). Finally, the change in R2 scores, ΔR2, is calculated (Figure 2A Step 3). This process is then repeated for all independent variables in the model, and the ΔR2 values are plotted in a heatmap for the variables of interest (Figure 2B-D).

Leave-one-covariate-out (LOCO) analysis for clinical-demographic features and reservoir characteristics while predicting immunophenotypes.

A. Explanation of leave-one-covariate-out (LOCO) analysis based on example of %CD4 T for clinicaldemographic features and reservoir characteristics while predicting immunophenotypes. Analysis was performed for all 133 immunophenotypes considered in the study. The top 10 biggest drops in adjusted R2 scores are reported for models that use total reservoir frequency (B.), intact reservoir frequency (C.), or percent intact (D.) as features in addition to clinical and demographic information. Participants with missing years of ART values are excluded from this analysis. The missing value of the CD4 nadir for one participant is imputed. In Figure S6, the LOCO analysis results are visualized for all 133 immunophe-

We applied the LOCO inference analysis to all 133 immunophenotype parameters (dependent variables). Three separate analyses were conducted for a set of independent variables consisting of age, sex, years of ART, CD4 nadir, recent CD4 account, years of HIV infection prior to ART (=NA, < 1, ≥ 1), and one of the three HIV reservoir metrics (total HIV DNA, intact HIV DNA, or percentage intact HIV DNA). Multicollinearity analysis confirmed that independent variables from this analysis were not correlated with each other to a degree that would interfere with model generation (Table S5). Each analysis produced a heatmap of 133 dependent immunophenotype variables, with visualization of the variables for each immunophenotype (Figure S6, Tables S6-S8). For each of these analyses, we displayed the top 10 variables with the largest ΔR2 value for LOCO inference incorporating either total (Figure 2B), intact (Figure 2C), or percentage intact (Figure 2D) HIV DNA.

The least-squares linear regression models were generally weakly explanatory (R2 values in range [-11.11%, 35.36%]) of the global variability in immunophenotypes (Tables S6-S8, Figure 2B-D, Figure S7). Across all 3 analyses, age and race emerged as an important covariate of several immunophenotypes related to T cell subsets, particularly the naïve and central memory compartments, as well as KLRG-1, a marker of immune senescence (Figure 2 and Figure S6). Interestingly, biological sex was a key variable for model prediction of NK cell NKG2A expression, and race was important for Tem CD8, KLRG1+/CD27+ CD8 frequency for all three models. Recent CD4 count was important for predicting the frequencies of HLA-DR-CD8 Tn cells and CD4, CD8 T cell frequencies, but more so for the models incorporating total HIV DNA and percent intact than the intact HIV DNA frequency (Figure 2A-C). Total HIV DNA frequency was an important model input for prediction of PD-1+ CD4 T naïve cells, HIV-specific IL-2+TNFα-IFNγ-CD107-CD4 T cells, and CD4/CD8 T cell frequencies, among others (Figure 2B, Figure S6). Total HIV DNA frequency and intact HIV DNA frequency were also important predictors of CD4 T effector memory HLA-DR+CD38+ expression (Figure 2B and 2C).

In general, total and intact HIV DNA were predictive of similar immune features (Figure S6, Tables S6-S8); however, total HIV DNA generally typically had greater magnitude ΔR2 than intact HIV DNA. The percentage of intact HIV DNA did not appear to contribute greatly to the predictive power of the model for the variance in the immunophenotypes examined (Figure S6C), most likely due to its relatively high correlation with years of ART Overall these findings further demonstrate the existence of a number of clinical parameters that are associated with immune signatures in PWH, including age, gender, and years of ART. Additionally, these results highlight key potential confounding variables that should be considered when interpreting the association of virological parameters with the immune system of PWH.

Receiver operator characteristic curve analysis demonstrates distinct immune parameters associated with intact and total reservoir frequencies

The complex relationship between the HIV reservoir and the immune system led us to evaluate if machine learning (ML) approaches that combine several parameters into models of the HIV reservoir could be a useful approach to understand this interaction. Before this could be attempted, we first identified a defined set of the most valuable variables from which ML models could be built. To achieve this, we first binarized the three HIV reservoir metrics (total HIV DNA frequency, intact HIV DNA frequency, %intact HIV DNA) into high (above median) or low (below median) reservoir groups and generated receiver operator (ROC) curves for the 144 clinical-demographic and immunophenotype parameters (Figure 3). The area under the curve (AUC) of the ROC curve indicates the ability of the feature (immune parameter or clinical/demographic information) to correctly identifying a participant as having qualitatively low or high total HIV DNA frequency (Figure 3A), intact HIV DNA frequency (Figure 3B), or percentage intact HIV DNA (Figure 3C). A model that randomly guesses high or low reservoir has an AUC of 0.5. In this analysis, 57 host variables had an AUC higher than 0.6 for one or more of total HIV DNA, intact HIV DNA, or percent intact DNA (Table S9). The most effective individual immune markers for classification of high versus low total HIV DNA frequency included %NKG2A+ CD4 T (AUC = 0.70), %PD-1+ Tn CD4 T (AUC = 0.68), and %CD38+/HLA-DR-CD8 T (AUC = 0.68). In contrast, the most effective markers for classifying based on the frequency of intact HIV DNA frequency included %CD127+ CD4 T (AUC = 0.71), %CD8 T (AUC = 0.66), and %Tn CD8 T (AUC = 0.65). Finally, for percentage intact HIV DNA, years of ART (AUC = 0.72), %CD107-IFNγ-IL-2+TNFα+ CD4 T (AUC = 0.65) and %KLRG1+/PD1+ CD4 T (AUC = 0.65) were the most effective (Figure 3). Similar to the findings in the Spearman correlations (Table 2) and LOCO inference analysis (Figure 2), there were more variables that had AUC values > 0.6 for total reservoir frequency (Table S9): 44 variables above this threshold for the total reservoir, but only 19 variables for the intact reservoir and 23 variables for percent intact. Overall, this approach allowed us to derive a ranked list of the most predictive immune parameters for each aspect of the HIV reservoir, and these highly ranked features were thus used for subsequent dimension reduction and ML modeling.

Receiver operating characteristic (ROC) curves identify PWH parameters that can classify reservoir characteristics.

For total reservoir frequency (A), intact reservoir frequency (B), and percent intact (C), ROC curves are plotted for all 144 immune markers, demographics, and clinical variables (shown in gray). Axes represent the true positive rate (TPR) and the false positive rate (FPR) for each variable for classifying study participants into low (below median) versus high (above median) reservoir frequency. ROC curves for ten variables with the highest AUC values are shown in color for each HIV reservoir characteristic. Striped black lines represent the ROC curves of a random model. For years of ART ROC curves, we exclude participants with missing years of ART val2u7es.

Dimension reduction reveals two clusters of PWH with distinct HIV reservoirs

To examine the overall structure of dataset we employed an unsupervised dimension reduction (DR) machine learning approach (PaCMAP (50)) using the 10 immune cell features with the highest AUC values for classifying participants based on total HIV DNA frequency, intact HIV DNA frequency, or percentage intact HIV DNA (Figure 3, Table S9). Interestingly, while no clear clustering was observed when we use intact reservoir frequency or percentage intact HIV DNA associated features, we observed two distinct clusters of PWH (cluster 1 and cluster 2, Figure 4A) when using total HIV DNA associated features. These clusters were of roughly equal size with 50 participants in cluster 1 and 65 in cluster 2. Projecting total reservoir frequency (above or below the median) onto the clustering plot, we observe a strong distinction between the two clusters in terms of total HIV DNA reservoir characteristics, (Figure 4B,C). Analysis of quantitative reservoir frequencies across clusters (Figure 4C-E) demonstrated that cluster 1 is characterized by a smaller total HIV reservoir frequency, but greater percentage of intact proviruses (Figure 4C-E). In contrast, cluster 2 is defined by a larger total reservoir frequency, but lower percentage of intact proviruses (Figure 4C-E). The frequency of intact proviruses also tended to be lower for cluster 1 but this difference was not significant. We also visualized the dataset using principal component analysis (PCA) (Figure S9), and observed that participants from each of the clusters identified by PaCMAP occupied different areas of the plot, although the separation was not as clear as with PaCMAP.

Dimension reduction reveals two major clusters of PWH with distinct immune systems and reservoirs.

A. PaCMAP was applied to the data using the ten immune cell features with the highest AUC values for classifying participants based on total reservoir frequency, and two clusters (cluster 1 and cluster 2) are identified. B. Same as A but data points are color-coded by total reservoir frequency (high=pink, low=gray). Total reservoir frequency (C.), intact reservoir frequency (D.), and percent intact (E.) is shown for participants within each cluster. F. Key immune cell features that distinguish cluster 1 from cluster 2 are identified by visualizing the features with the highest AUC values with respect to classifying cohort participants based on cluster membership. Axes represent the true positive rate (TPR) and the false positive rate (FPR) for each variable. Immune markers and clinical-demographic features are shown for each cluster in Figure S8.

To gain insight into the immune cell features that distinguish these two clusters, we generated ROC curves for the 133 immune parameters based on their ability to identify membership in cluster 1 versus cluster 2 for each participant (Figure 4F). Many features that distinguished the clusters overlapped with the features selected for the clustering analysis, including expression patterns of CD38, HLA-DR, KLRG-1 and PD-1 on CD4 and CD8 T cells and CD4 T naïve cell frequencies. However, some novel features emerged, including CD4 KLRG1-CD27+ frequency, CD8 KLRG1-PD1-frequency, and CD8 Tn frequency (Figure 4F and Figure S8). Cluster 2, which had a higher total and intact absolute reservoir frequency but lower percentage of intact proviruses, was associated with a lower fraction of naïve CD4 and CD8 T cells. Additionally, cluster 2 had generally higher expression of immune exhaustion markers, including KLRG1 and PD-1 (Figure S8). These data indicate that cluster membership is highly associated with the overall level of immune activation and exhaustion for the individual. Notably, with regard to other clinical variables, cluster 2 had an older median age, longer time on ART, lower CD4 nadir, and lower current CD4 T cell count (Figure S8). We have previous shown that cannabis use is associated with lower levels of immune activation markers in PWH, despite having minimal impact on the size of the HIV reservoir (51). Interestingly, a subset of participants in this study had clinical data regarding cannabis (CB) use. CB users were enriched in cluster 1, which had lower levels of immune exhaustion markers (Figure S8). Overall, these data demonstrate that, within the cohort of PWH on long-term ART, there were two major clusters of participants with distinct reservoir characteristics and immunophenotypes.

Decision tree visualization of PWH with respect to reservoir characteristics

Since the interaction of the immune system and the HIV reservoir is multifactorial, we hypothesized that models that consider multiple parameters simultaneously could more accurately describe the overall dataset, and provide insights regarding the biology of the HIV reservoir and the host immune system. To accomplish this, we employed a decision tree approach to visualize combinations of variables that classify participants as having high (above median) or low (below median) reservoir frequency. Compared to the dimension reduction technique, decision tree visualization is an interpretable supervised approach and does not require post-hoc analysis.

We first selected 35 variables with the highest ROC AUC values for either total reservoir frequency, intact reservoir frequency, or percentage intact HIV DNA to be considered for model generation (Figure 4). Using these parameters, we fitted Generalized and Scalable Optimal Sparse Decision Trees (GOSDT) (52) to the data in order to classify high versus low total HIV DNA, intact HIV DNA, or percentage intact HIV DNA. We required the trees to achieve at least 80% accuracy for classifying PWH in the cohort, as well as have at least five PWH in each leaf. Since these trees are based on the entire dataset, these models are thus descriptive rather than predictive.

For the total HIV DNA frequency decision tree, only four immune variables were required to accurately describe high versus low total HIV DNA status (Figure 5A, Figure S10): CD8 T cell frequency, CD4 nadir, %CD38+HLA-DR-CD8 T cells, and %NKG2A+ CD4 T cells. The tree divided the cohort into five subgroups (leaves), among which three have high total reservoir frequency and two have low total reservoir frequency. Comparing the labels provided by GOSDT model with the actual data, the tree achieved 83.5% accuracy (i.e. misclassifying 19 PWH among the overall cohort of 115 PWH). Notably, when we combined all samples from “high total reservoir” leaves and all samples from “low total reservoir” leaves, we observed a significant difference in the actual median total reservoir frequencies for these two groups (266/M for low total and 1288.5/M for high total, Mann Whitney U test p-value is 3.56e-13, Figure 5B).

Decision tree visualization of the association of immune cell subsets with reservoir characteristics.

(A, C, E) Host variables (immune cell frequencies, demographic, and clinical information) were used to visualize the PWH dataset using the optimal sparse decision trees algorithm GOSDT. The overall set of PWH was classified as likely having high (above median, orange “leaves”) or low (below median, blue “leaves”) total reservoir frequency (A), intact reservoir frequency (C), and percent intact (E). In each leaf, “med” denotes the median HIV characteristic of PWH, N is the number of PWH in the leaf, and MN is the number of misslabeled PWH. (B, D, F) PWH in model leaves associated with high (orange) or low (blue) reservoir frequency characteristics were aggregated and a Mann-Whitney U test was performed to determine statistical significance between the actual total reservoir frequency of the “high” and “low” groups for total reservoir frequency (B), intact reservoir frequency (D), and percent intact (F). For the percent intact tree we exclude participants with missing values of years of ART. For total and intact reservoir frequency, missing values of years of ART were imputed, however, since the trees do not use this variable, imputations do not influence results. Visualization trees are explained with sets of rules in Supplemental Materials: Figure S10 for total reservoir frequency, Figure S11 for intact reservoir frequency, and Figure S12 for percent intact.

We then repeated this approach for classifying the cohort participants with respect to intact reservoir frequency (Figure 5C, Figure S11). Despite the overall lower correlation between the immune cell phenotypes and the intact reservoir size, this tree nevertheless achieved 82.6% accuracy. The tree had six leaves (three with low intact reservoir, three with high intact reservoir) and relied on %Tn CD8 T cells, %CD107a-IFNγ+IL-2+TNFα-CD4 T cells, %CD127+ CD4 T cells, %CD38-HLA-DR+ Tcm CD8 T cells, and CD4 Nadir. When we combined the ‘high intact’ leaves together and the ‘low intact’ leaves together, we observed a significant difference in median intact reservoir frequency between the groups (29.5/M for low intact and 101/M for high intact, Mann Whitney U test p-value is 9.59e-08, Figure 5D).

Finally, a third iteration of GOSDT generation was performed, classifying the cohort participants with respect to percentage intact of total proviruses (Figure 5E, Figure S12). This tree achieved 82.4% accuracy and relied on years of ART treatment, %KLRG1+CD27-CD4 T cells, %CD127+ CD4 T cells, and recent clinical CD4 count. This tree has two leaves with low percent intact reservoir and three leaves with high percent intact reservoir. We observed a significant difference in the actual percent intact values between the model generated high percent intact and low percent intact leaves (median 5.7% for low percent intact leaves and median 15.3% for high percent intact leaves: Mann Whitney U test, p-value is 5.5e-10, Figure 5F).

Collectively, these decision trees highlight combinations of immune parameters that can accurately describe qualitatively high versus low total, intact, and percentage intact HIV DNA in a cohort of n = 115 PWH. This visualization serves as a basis for mechanistic hypotheses about the interactions of the immune system and HIV reservoir during long-term ART.

Machine learning algorithms identify immune feature combinations that predict high versus low reservoir metrics with ∼70% test accuracy

Although clustering and decision tree analysis permit visualization and understanding of global structures within a dataset, we were curious if combinations of immune and clinical-demographic parameters could actually accurately predict, rather than only visualize, the size of the HIV reservoir. We considered five machine learning algorithms including Logistic Regression with L2 regularization (LR), CART, Support Vector Machines with RBF kernel (SVM), Random Forest (RF), and Gradient Boosted Trees (GBT). Initially, we attempted to predict reservoir frequency as a continuous variable, but found that models trained for this purpose performed poorly and tended to overfit to training sets (Figure S13). Thus, we focused our efforts on predicting high (above median) versus low (below median) reservoir frequency. For reservoir characteristics, we measured accuracy of the models over 10 random splits of our data into training and test sets (Figure 6A and Table S10). Overall, we found that for total reservoir frequency, Logistic Regression achieved highest mean classification accuracy (69.31%) in test data. The accuracy of these models is likely limited by the sample size (n=115) and noise in the data, as evidenced by lower R2 score of machine learning models for non-binarized HIV characteristics (Figure S13).

Predicting HIV reservoir characteristics with machine learning.

Average training and test accuracies over ten training and test data splits for Random Forest (RF), Gradient Boosted Trees (GBT), Support Vector Machines with RBF kernel (SVM), Logistic Regression (LR), and CART models for total reservoir frequency (A), intact reservoir frequency (C), and percent intact (E) are shown. For one split of training and test sets, logistic regression models are visualized for total reservoir (B), intact reservoir (D), and percent intact (F). On the y-axis we show variables used by the model, while the x-axis displays coefficient values for individual variables used by models. For percent intact models we exclude participants with missing values of years of ART. For total and intact reservoir frequency, missing values of years of ART were imputed. The missing value of the CD4 nadir for one participant was imputed using the MICE algorithm.

To examine the contribution of individual immune features to model performance, we examined lo-gistic regression coefficients for each immune cell variable in the model. Since Logistic Regression coefficients are associated with expected change in log odds (based on loge), we can think about the coefficient β for variable X in the following way: increasing variable X by one unit multiplies the odds of high reservoir frequency (probability that the reservoir size is high divided by the probability that the reservoir frequency is low) by eβ. In Figure 6B we visualize the logistic regression model for one data split among ten we considered for total reservoir frequency. For this split we observe that higher values of %NKG2A+ CD4 T, %PD-1+ Tn CD4 T, and %Tcm CD8 T are associated with a increased probability of total reservoir frequency being high. On the other hand, an increase in %Tn CD4 T decreases the odds of high reservoir frequency. The model visualized in Figure 6B achieved 75.86% training and 75% test accuracy.

Similarly, logistic regression models performed best for predicting intact reservoir frequency compared to other methods (average 65.17% test accuracy, see Figure 6C). In Figure 6D we visualize the Logistic Regression model for one fixed data split. In this model we observed that higher values of %CD107a-IFNγ+IL-2+TNFα-CD4 T and %CD127+ CD4 T, and % CD4 T are associated with lower probability of intact reservoir frequency being high, while higher values of %KLRG1+CD27+ CD8 T are associated with increased probability of intact reservoir frequency being high.

For analysis of percentage intact HIV DNA, we display training and test accuracy values in Figure 6E. For the visualized model for one data split (Figure 6F), an increase in %CD127+ CD4 T, years of ART, %KLRG1+PD-1+ CD4 T, and %Tem CD4 T leads to a lower probability of percent intact HIV DNA being high.

Overall, these analyses demonstrate that we can use machine learning tools to construct models that can predict with approximately 70 percent accuracy whether a given PWH has qualitatively low or high total HIV DNA frequency, intact HIV DNA frequency or percentage of HIV DNA. Further studies with larger cohorts will likely improve the accuracy of these models.

Discussion

In this cross-sectional study of peripheral blood samples from a well-characterized clinical cohort of 115 PWH on long-term ART, we used machine learning analysis of IPDA and high parameter flow cytometry to better understand the associations between the HIV reservoir and a broad range of immunophenotypes. Using these approaches, specific HIV reservoir features (total HIV DNA frequency, intact HIV DNA frequency, percentage intact HIV DNA) and clinical-demographic variables (age, sex, CD4 nadir, recent CD4 count, years on ART, years of HIV infection prior to initiating ART) were identified that predict specific immunophenotypes in PWH on long-term ART. Distinct host immune correlates of intact versus total HIV DNA frequency were also identified. Further, combinations of immune parameters that classified and predicted whether a given PWH has high or low total, intact, or percentage intact HIV DNA were defined. Collectively, this study confirms existing knowledge about HIV reservoir biology and identifies novel associations for further investigation. Additionally, from a methodological perspective, this work also demonstrates the utility of specific data science and machine learning approaches for HIV reservoir studies. Development and application of these analytic approaches are important to enable biological insights from high-parameter studies of the HIV reservoir and other complex biology, particularly with the advent of high-dimensional techniques such as CYTOF and scRNAseq.

Previous studies have sought to identify immune and other correlates of the total HIV DNA reservoir (21,26–44). These studies have identified several activation/exhaustion markers that correlate with total HIV DNA, including the CD4/CD8 ratio, T cell expression of HLA-DR, CD38, PD-1, CTLA-4, and LAG-3, and NK cell KLRG1 expression. However, not all studies have found associations of total HIV DNA and immune subsets, particularly for CD4 T cell HLA-DR and CD38 expression (39, 53–55) during longterm ART. There is also growing appreciation for additional factors that correlate with the HIV reservoir. Specifically, the size of the reservoir has been linked to the abundance of specific bacteria species in the intestinal microbiome (56). Also, genetic studies have indicated the existence of single nucleotide polymorphisms (SNPs) that are associated with reservoir size, including SNPs in MX1 and IRF7 (57–59). The correlative results from this present study corroborate many of these studies, and provide additional insights. A number of correlations between total HIV DNA and immune activation and exhaustion were identified. There were weak to moderate positive correlations observed between total HIV DNA frequency and CD4 T cell expression of NKG2A, PD-1, KLRG1 and HLA-DR as well as CD8 T cell expression of PD-1 and CCR7. Total HIV DNA frequency was inversely correlated with CD8 T cell expression of CD38, CD4 T cell expression of CD27, the frequency of naïve CD4 and CD8 T cells, and the CD4/CD8 ratio. Overall our results support the hypothesis that larger HIV reservoir size is associated with with increased frequencies of cells with an activated or exhausted phenotype, with depletion of naive cells and with an increased proportion of central memory CD8 T cells. The correlation of NKG2A expression in CD4 T cells with the total reservoir size represents a novel observation that merits further investigation. Only a very small percentage of CD4 T cells express NKG2A (less at 1 percent), and the biological significance for the correlation of these cells with the HIV reservoir is unknown. NGK2A is known to function as an inhibitory receptor for NK cells and for CD8 T cells, but its role in CD4 T cells is less clear (60).

When we examined immune correlates of intact HIV DNA measured by the IPDA, many of the same relationships as described above were noted, but with decreased magnitude and often lacking statistical significance. This is consistent with a recent study using the IPDA that did not find associations of intact HIV DNA with CD4 T cell HLA-DR and CD38 co-expression (61), as well as a study using near-full-length viral sequencing in single cells that found only slight enrichment of HLA-DR or PD-1 in cells with genetically intact, inducible proviruses (36). The stronger relationship between the immune cell phenotypes and the total reservoir, rather than the intact reservoir, should be investigated further. We speculate that this association could be driven by the more numerous defective proviruses promoting inflammation or immune activation through innate immune sensing pathways, or a greater level of transcriptional activity for defective proviruses. This is certainly plausible giving the growing appreciation for the immunogenicity and high frequency of defective proviruses within the HIV reservoir (62–65). It is important to note, however, that it is also possible that the stronger association of total proviral burden with immunophenotype reflects persistent damage caused to the immune system before therapy, and that this association between the total reservoir and immune cells may not result from ongoing interactions between HIV and the immune system during therapy.

Uniquely, CD127 expression on CD4 T cells was significantly associated with the intact reservoir frequency but not the total reservoir. Specifically, the frequency of CD127-CD4 T cells was positively correlated with the frequency of intact proviruses. CD127 is the alpha-chain IL-7 receptor and is expressed on long-lived memory T cells that are depleted during untreated HIV infection. Post-ART, the frequency of CD127+ memory CD4 subsets in PWH increases over time, albeit slowly (11). Interestingly CD127+ cells are enriched in latent HIV infection in tissues (66), and IL-7 is associated with slower natural reservoir decay (27). Furthermore, IL-7 signaling is known to drive expression of VLA-4, which was recently reported to highly enrich for genetically intact, inducible proviruses (36). This observational finding that intact but not total HIV DNA inversely correlates with CD127 expression on CD4 T cells requires further investigation.

Dimension reduction machine learning approaches identified two robust clusters of PWH when using total HIV DNA reservoir-associated immune cell frequencies. These clusters exhibited distinct reservoir characteristics, with one cluster being enriched with participants with a larger total reservoir size, a lower percentage of intact viruses, a lower frequency of naïve T cells and elevated expression of activation and exhaustion markers, while the other contained PWH with smaller total reservoirs, higher percentage of intact viruses, higher frequency of naïve T cells and lower expression of activation and exhaustion markers. The cluster membership was notably associated with age and time on therapy. The existence of two distinct clusters of PWH with different immune features and reservoir characteristics could have implications for HIV cure strategies. These two clusters are also interesting in the context of a previous study examining reservoir characteristics which proposed two major types of viral reservoir within PWH: one type with smaller reservoirs that were enriched in Tcm cells and associated with lower levels of Ki67+ cells and immune activation, and another type with larger reservoirs that were enriched in Ttm cells and associated with higher levels of Ki67+ cells and immune activation (27). It is possible that these two proposed types of reservoirs represent the two clusters we observe using immune markers and dimension reduction, although we were unable to examine the frequency of HIV proviruses specifically within Tcm and Ttm cells in this study.

In addition to unsupervised dimension reduction approaches, we were able to construct simple, in-terpretable decision trees that describe our cohort of PWH with more than 80% accuracy with respect to the size of the reservoir. Furthermore, when we constructed decision trees to classify PWH separately based on total reservoir frequency, intact reservoir frequency and percentage intact, distinct sets of features were important (though some were the same) for each of these models. The decision tree analyses highlight complex immune parameter combinations that serve as a basis for unique hypotheses about HIV reservoir biology. One striking example of machine learning approaches identifying potential reservoir biology is the percentage intact HIV DNA decision tree. As expected, years of ART was critical to predict the percentage of intact proviruses, but CD4 T cell CD127 expression, CD4 T cell KLRG1 expression and recent CD4 count also played a significant role.

Logistic regression models that we trained to classify high versus low total reservoir, intact reservoir, and percentage intact were able to predict out-of-sample with ∼70% accuracy on average. Each of these models relied on five features only and were simple enough for us to visualize and assess how important each variable is for every characteristic of reservoir size. These models corroborated the potential biological relationships identified in the decision tree analysis, and confirmed that these variables could actually be used to predict out-of-sample whether total, intact or percentage intact HIV DNA was qualitatively high or low. The success of our predictive models was likely limited by the relative small size of the datasets used to train and test them (n=115). As such, larger datasets will be required to generate more accurate models that can be used for reliable prediction of reservoir characteristics. Additionally, it is likely that unmeasured factors and biological noise also contribute to reservoir size.

Our findings should be considered in the light of some inherent limitations and caveats. This study is cross-sectional in nature and is primarily observational, so the findings should be interpreted with caution. As with any observational study, confounding bias may influence our results, particularly for correlative analyses. The associations between the HIV reservoir and the immune system we observed were fairly weak, despite being statistically significant. Similarly, the predictive ability of our ML models was modest (70 percent) for classifying PWH as having an above or below median reservoir. As mentioned above, our cohort is likely too small to generate accurate predictive ML models, and consists of mostly male participants, which could bias the results. We also cannot rule out the influence of additional unmeasured confounding factors. In particular, the length of untreated infection prior to ART could be a major contributor to the immune signatures we observe. In this study, the participants had all been treating during chronic infection, and different associations might be observed in PWH who are treated during acute infection. Another limitation to this study is that our analyses only considered cells and proviruses in the peripheral blood, and this may not represent HIV reservoirs and immune cells located in tissues known to harbor latent HIV such as gut, spleen and brain. Our analysis of surface protein phenotypes also does not detect potentially important associations between the reservoir and transcriptional pathways within these cell types. Future work using single cell RNAseq from a similar cohort may help to reveal deeper layers of associations that are invisible to the current analysis. As is always the case with correlation studies, its is important to be aware that these associations are only correlations and do not necessarily represent functional or mechanistic relationships. Even so, these correlations are still useful as a source of novel hypotheses that could be tested in subsequent studies.

Nevertheless, our study does have some strengths compared to previous work in this area, specifically a highly detailed flow panel, allowing us to quantify numerous immune subsets with a high degree of resolution. Also, the use of the intact proviral DNA assay allows us to separately examine the association of the immune system with intact and total viral DNA. It is however, important to note that IPDA does not measure all proviral sequences in PWH and is likely failing to identify some sequences due to proviral polymorphisms or dual deletions in the two IPDA amplicons. Also, a minor fraction of IPDA+ proviruses are actually defective due to the limited specificity of using two short amplicons to select for full-length intact proviruses (46, 67).

Overall, these findings suggest a complex concert of immune recovery, HIV reservoir dynamics, and intrinsic host factors (age, biological sex) that shapes host immunophenotype, even after years of ART. Mechanistic work will be needed to fully dissect the dynamic relationship between the immune system and the reservoir. Nevertheless, these findings support a model in which ongoing interaction between the HIV reservoir and the host immune cells continue to drive an association, albeit a relatively minor one, with persistent CD4 T cell activation during long-term ART. Notably, both intact and defective proviruses appear to contribute to this immune signature of HIV latency during long-term ART. This study identifies several areas for future investigation. The question of whether the HIV reservoir directly drives persistent immune activation could be addressed directly by specifically suppressing viral transcription during therapy and examining the impact on the immune system. The stronger association of the intact reservoir with CD127 expression in CD4 T cells is an intriguing observation that warrants further validation and explanation. This observation could, for example, be explained by a model in which the intact reservoir is preferentially located in cells with a distinct pattern of CD127 expression. These ideas, along with other future studies exploring the interplay of the host immune system and HIV reservoir might leverage new high-dimensional technologies such as scRNAseq and mass cytometry. The machine learning approaches described and applied in this study may be particularly useful for gaining biological insights into these high parameter datasets.

Methods

Cohort and sample collection

115 people with HIV (PWH) were recruited from two clinical sites. 66 PWH were recruited at Duke University medical center and 49 PWH at UNC Chapel Hill. Peripheral blood was collected and peripheral blood mononuclear cells (PBMCs) were obtained by Ficoll separation, then frozen in 90% fetal bovine serum 10% dimethylsulfoxide (DMSO) (51, 68–70).

Flow cytometry

Cell preparation and staining followed previously described methods (71). All antibodies were titrated to optimize signal-to-noise ratio on PBMCs prior to use, assuming a 50μL staining volume. Cell viability was examined using Zombie-NIR Fixable Viability Dye (0.4μL per 50μL staining volume; Biolegend). Antibodies used for surface staining were as follows: KLRG1-BV421 (SA231A2 clone, Biolegend, CD45RA-PacBlue (H100 clone, BL), CD8-BV570 (RPA-T8 clone, Biolegend), CD127-BV605 (A019D5 clone, Biolegend), CD56-BV650 (5.1H11 clone, Biolegend), CCR7-BV711 (G043H7 clone, Biolegend), CD27-BV750 (O323 clone, Biolegend), PD1-VioBright515 (REA165 clone, Miltenyi), NKG2A-PE-Vio615 (REA110 clone, Miltenyi), CD16-PerCP-Cy5.5 (33G8 clone, Biolegend), CD38-PCPeF710HB7 clone, TF), CD14-SparkNIR685 (63D3 clone, Biolegend), CD19-SparkNIR685 (HIB19 clone, Biolegend), HLA-DR-APC-F750 (L243 clone, Biolegend). Antibodies used for intracellular staining were as follows: CD3-BV480 (UCHT1 clone, BD), CD4-PerCP (L200 clone, BD), IFN-g-PE-Cy7 (4S.B3 clone, Biolegend), IL-2-APC (MQ1-17H12 clone, Biolegend), TNFα-AF700 (Mab11 clone, Biolegend). For HIV peptide stimulation, the cells were incubated with peptide mixes from Gag, Pol and Env proteins at 0.2µg/mL in the presence of brefeldin A and monesin for 6 hrs. Samples were analyzed using a Cytek Aurora spectral flow cytometer. Flow cytometry analysis was performed in FlowJo v10.8 software.

Intact Proviral DNA Assay (IPDA)

Cryopreserved samples of PBMCs from each study participant were viably thawed. A portion was used for immunophenotyping as described above, and the remainder were subjected to total CD4 T cell negative selection with the StemCell Technologies EasySepTM Human CD4+ T Cell Enrichment Kit (Cat# 19052). CD4 T cell DNA was extracted using the QIAamp DNA Mini Kit and quantified on a NanoDrop 1000 (Thermo Fisher Scientific). IPDA was performed as originally described (46), with a validated PCR annealing temperature modification to increase signal to noise ratio (70). Gating for positive droplets was set using negative (DNA elution buffer and HIV-seronegative CD4 T-cell DNA), and positive (Integrated DNA Technologies gblock amplicon) positive control wells processed in parallel. DNA shearing index values were similar to those reported previously (median, 0.33; interquartile range [IQR], 0.31– 0.34) (68–70). A median of 1.04x106(Q1 − 8.71x105, Q3 − 1.16x106) cell equivalents were assessed for each donor.

Statistics

Variable data analysis

To create binarized labels that represent reservoir characteristics, we split the data at the median, which is 553/M for total reservoir frequency, 53/M for intact reservoir frequency, and 8.64 percent for percent intact. For every variable (immune cell frequency, demographics, and clinical information), we computed Spearman correlation and an Area Under the Curve (AUC) value. To compute the AUC value, we first created a Receiver Operating Characteristic (ROC) curve by plotting the true positive rate versus the true negative rate for every cell subset frequency. AUC is then computed as the area under the ROC curve using the trapezoidal rule.

To assess the importance of clinical, demographic information and HIV reservoir characteristics in describing the immune markers, we performed a leave-one-covariate-out (LOCO) analysis. We trained a linear regression model (M1) with intercept based on all variables. We removed each variable one at a time and trained a new model without this variable. Then we computed the difference in R2 score before and after dropping the variable. Variance Inflation Factors (VIFs) for age, sex, years of ART, CD4 nadir, recent CD4 account, and years of HIV infection prior to HIV infection were less than 2, indicating an acceptable level of correlation between these independent variables (72). However, higher VIF values were observed for intact and total HIV DNA, therefore separate LOCO inference analyses were conducted in order to avoid artificial fluctuations in model fit due to multicollinearity.

Data visualization

We used PaCMAP (50) to reduce the dimensionality of the dataset to a 2-dimensional space. Due to the relatively small sample size of our dataset (n=115), we set the number of neighbors to 5. Additionally, we used GOSDT (52) which computes sparse optimal trees to identify patterns in the dataset. For all the visualization trees we set the depth budget parameter to 4 and the regularization parameter to 0.02.

Training and generalization

Our training procedure for classification is described in Algorithm 1 in Supplemental Materials, which is a standard machine learning pipeline with an added search for the number of variables. For the full hyperparameters list see Table S8. We normalized continuous variables to have values between 0 and 1. For logistics regression models, we set the regularization parameter of ℓ − 2 regularization based on cross-validation. The values are 10 for total reservoir frequency and percent intact and 100 for intact reservoir frequency.

Study approval

Written informed consent was obtained for all study participants. The study design was reviewed and approved by IRB for both Duke University and UNC Chapel Hill.